### 2014-09-09 article/article.tex
latex outline article, fixes #2

% !TEX TS-program = pdflatex
% !TEX encoding = UTF-8 Unicode

% This is a simple template for a LaTeX document using the "article" class.
% See "book", "report", "letter" for other types of document.

\documentclass[11pt]{article} % use larger type; default would be 10pt

\usepackage[utf8]{inputenc} % set input encoding (not needed with XeLaTeX)

%%% Examples of Article customizations
% These packages are optional, depending whether you want the features they provide.
% See the LaTeX Companion or other references for full information.

%%% PAGE DIMENSIONS
\usepackage{geometry} % to change the page dimensions
\geometry{a4paper} % or letterpaper (US) or a5paper or....
% \geometry{margin=2in} % for example, change the margins to 2 inches all round
% \geometry{landscape} % set up the page for landscape
%   read geometry.pdf for detailed page layout information

\usepackage{graphicx} % support the \includegraphics command and options

% \usepackage[parfill]{parskip} % Activate to begin paragraphs with an empty line rather than an indent

%%% PACKAGES
\usepackage{booktabs} % for much better looking tables
\usepackage{array} % for better arrays (eg matrices) in maths
\usepackage{paralist} % very flexible & customisable lists (eg. enumerate/itemize, etc.)
\usepackage{verbatim} % adds environment for commenting out blocks of text & for better verbatim
\usepackage{subfig} % make it possible to include more than one captioned figure/table in a single float
\usepackage{amsmath} 
% These packages are all incorporated in the memoir class to one degree or another...

%%% HEADERS & FOOTERS
\usepackage{fancyhdr} % This should be set AFTER setting up the page geometry
\pagestyle{fancy} % options: empty , plain , fancy
\renewcommand{\headrulewidth}{0pt} % customise the layout...
\lhead{}\chead{}\rhead{}
\lfoot{}\cfoot{\thepage}\rfoot{}

%%% SECTION TITLE APPEARANCE
\usepackage{sectsty}
\allsectionsfont{\sffamily\mdseries\upshape} % (See the fntguide.pdf for font help)
% (This matches ConTeXt defaults)

%%% ToC (table of contents) APPEARANCE
\usepackage[nottoc,notlof,notlot]{tocbibind} % Put the bibliography in the ToC
\usepackage[titles,subfigure]{tocloft} % Alter the style of the Table of Contents
\renewcommand{\cftsecfont}{\rmfamily\mdseries\upshape}
\renewcommand{\cftsecpagefont}{\rmfamily\mdseries\upshape} % No bold!

%%% END Article customizations

%%% The "real" document content comes below...

\title{N-Body problem}
\author{Marcel Boersma\\Shabaz Sultan}
%\date{} % Activate to display a given date or no date (if empty),
         % otherwise the current date is printed

\begin{document}
\maketitle

\section{Introduction (phenomena)}
\section{Literature Study}
\section{Methodology}
\subsection{Discrete Approximation}
\subsection{Algorithms}
\subsection{Implementation}
\section{Experimental Results}
\section{Conclusions}

\end{document}

### 2014-09-13 article/article.tex
Changes ignore file and adds subheader for the literature study

@@
 \section{Introduction (phenomena)}
+A world famous problem is the two body problem, most commonly known as the sun - earth problem.
+
 \section{Literature Study}
+background etc
+\subsection{Newton's law}
+\begin{equation}
+ F = G\frac{m_1m_2}{r^2}
+\end{equation}
+With G as a gravitational constant and F as the force in Newton and $m_1,m_2$ the mass of the two systems in kg and $r$ is the distance between the center of the two bodies (see figure xx). 
+Keplers problem? 
+\subsection{Kepler}
+Kepler problem, special case the Kepler orbit problem
+
 \section{Methodology}
+Newton's law, Kepler orbit for next position.
+
 \subsection{Discrete Approximation}
+Approximation of Newton's law, Kepler orbit. Possible accuracy problems, e.g. sensitivity analysis.
+
 \subsection{Algorithms}
+
 \subsection{Implementation}

### 2014-09-16 article/article.tex
Adds more mathematical/physics background information

@@
 \section{Methodology}
-Newton's law, Kepler orbit for next position.
+Assume that we have an initial velocity vector $\overrightarrow{v_t}$ for the body t with speed in both x and y direction.
+\begin{equation}
+	\overrightarrow{v_t} = \begin{bmatrix}
+								x_t \\
+								y_t
+							\end{bmatrix}
+\end{equation}
+For example we have a two body system with $p,c$ with an initial velocity vector, respectively, then
+\begin{equation}
+	v_p=\begin{bmatrix} v_x^p \\ v_y^p \end{bmatrix}, v_c=\begin{bmatrix} v_x^c \\ v_y^c \end{bmatrix}
+\end{equation}
+We assume that the whole system has a constant velocity, in our case we assume zero velocity, hence the next equation must be equal to zero
+\begin{equation}
+	V = \frac{P}{M}	
+\end{equation}
+with $P$ as the total momentum and $M$ as the total mass of the system.
+We know that the total momentum is the sum of all forces acted up on and the mass is a constant.
+Therefore, our zero velocity implies that the total momentum must be equal to zero.
+Or more formally
+\begin{equation}
+	P = m_p\overrightarrow{v_p} + m_c\overrightarrow{v_c} = 0.
+\end{equation}
+As illustrated in figure~\ref{XXXX}(simple two body system with vectors drawn) we can see that each body has several forces acted upon.
+Our objective is the determine the path of body $p,c$ and as you can see we have a force vector $F_x$ pulling the body $X$ to the center and also our velocity vector $v_x$.
+The product of those vectors gives us the directional vector of our body $X$.
+Due to the fact that the two bodies are the only objects in the system acting on eachother we can determine the radius and period of our movement around a (virtual) center of mass.
+We know for a fact that the two bodies must be the opposites of eachother such that the total momentum is equal to zero.
+The center of mass can be calculated in reference of one of our bodies.
+\begin{equation}
+	(m_p + m_c)\overrightarrow{r_{m}} = m_p\overrightarrow{r_p} + m_c\overrightarrow{r_c} 
+\end{equation}
+With $r_m$ being the vector to the center of mass, the we can rewrite this to calculate $r_m$
+\begin{equation}
+	\overrightarrow{r_m} = \frac{m_p\overrightarrow{r_p} + m_c\overrightarrow{r_c}}{(m_p + m_c)}
+\end{equation}
+and we will calculate the $\overrightarrow{r_m}$ with as reference point our body $C$, then the $\overrightarrow{r_c}=0$ because the distance between our body $C$ and $C$ is zero, and the $\overrightarrow{r_p}=R$ with $R$ as the total distance between the two bodies.
+Then $\overrightarrow{r_m}$ becomes
+\begin{equation}
+	\overrightarrow{r_m} = \frac{m_p}{M}\overrightarrow{R}
+\end{equation}
 

### 2014-09-18 code/main.py
basic opengl python program with 3d particles, with readme describing the program

@@ -1,0 +1,96 @@
+import pygame
+import sys
+import random
+import math
+from OpenGL.GL import *
+from OpenGL.GLU import *
+
+# general OpenGL initialization
+def init_opengl(width, height):
+    glViewport(0, 0, width, height)
+    glMatrixMode(GL_PROJECTION)
+    glLoadIdentity()
+    gluPerspective(90.0, float(width)/height, 0.1, 100.0)
+    glMatrixMode(GL_MODELVIEW)
+    glLoadIdentity()
+    glEnable(GL_DEPTH_TEST)
+    glShadeModel(GL_SMOOTH)
+    glClearColor(0.1, 0.1, 0.15, 0.0)
+    glPointSize(2)
+    
+
+
+class Particle:
+    def __init__(self):
+        # generate random start position
+        self.x = random.random() * 10 - 5
+        self.y = random.random() * 10 - 5
+        self.z = random.random() * -10
+
+        # generate random start velocity
+        magnitude = 2
+        self.vx = (random.random() - 0.5)*magnitude
+        self.vy = (random.random() - 0.5)*magnitude
+        self.vz = (random.random() - 0.5)*magnitude
+
+
+    def render(self):
+        glColor3f(1, 1, 1)
+        glVertex3f(self.x, self.y, self.z)
+
+    def update(self):
+        dt = 0.1
+        gravity_strength = 0.5
+
+        # create acceleration vector by taking difference from particle's 
+        # position and (0,0,-5), which we're using as the source of gravity
+        ax = - self.x * dt * gravity_strength
+        ay = - self.y * dt * gravity_strength
+        az = (-5 - self.z) * dt * gravity_strength
+
+        # update velocity using acceleration vector
+        self.vx += ax * dt
+        self.vy += ay * dt
+        self.vz += az * dt
+
+        # update position using velocity vector
+        self.x += self.vx * dt
+        self.y += self.vy * dt
+        self.z += self.vz * dt
+        
+
+def main():
+    width = 800
+    height = 600
+    size = width, height
+
+    pygame.init()
+    screen = pygame.display.set_mode(size, pygame.DOUBLEBUF | pygame.OPENGL)
+    init_opengl(width,height)
+
+    # generate some particles
+    particles = []
+    for i in xrange(1000):
+        particles.append(Particle())
+
+    while True:
+        for event in pygame.event.get():
+            if event.type == pygame.QUIT: sys.exit()
+
[…]

### 2014-09-21 code/2body.py
Adds nbody simulation

@@ -1,0 +1,133 @@
+import pygame
+import sys
+import random
+import math
+from OpenGL.GL import *
+from OpenGL.GLU import *
+
+# general OpenGL initialization
+def init_opengl(width, height):
+    glViewport(0, 0, width, height)
+    glMatrixMode(GL_PROJECTION)
+    glLoadIdentity()
+    gluPerspective(90.0, float(width)/height, 0.1, 100.0)
+    glMatrixMode(GL_MODELVIEW)
+    glLoadIdentity()
+    glEnable(GL_DEPTH_TEST)
+    glShadeModel(GL_SMOOTH)
+    glClearColor(0.1, 0.1, 0.15, 0.0)
+    glPointSize(2)
+    
+
+
+class Particle:
+    def __init__(self):
+        # generate random start position
+        self.x = random.random() * 2 - 5
+        self.y = random.random() * 2 - 5
+        self.z = random.random() * -10
+
+        # generate random start velocity
+        magnitude = 0.1
+        self.vx = (random.random() - 0.5)*magnitude
+        self.vy = (random.random() - 0.5)*magnitude
+        self.vz = (random.random() - 0.5)*magnitude
+
+        self.mass = (random.random())*1000000;
+        self.ax = 0;
+        self.ay = 0;
+        self.az = 0;
+
+
+
+    def render(self):
+        #glColor3f(1,1,1)
+        #glColor3f(math.sqrt(math.pow(self.vx,2)), math.sqrt(math.pow(self.vy,2)), math.sqrt(math.pow(self.vz,2)))
+        #glVertex3f(self.x, self.y, self.z)
+        glPointSize((self.mass/1000000)*10)
+        glBegin(GL_POINTS)
+        glColor3f(1,1,1)
+        glVertex3f(self.x, self.y, -10);
+        glEnd()
+
+    def mass(self):
+        return self.mass;
+
+    def update(self, particles):
+        dt = 0.1
+        gravity_strength = 0.5
+
+        self.acceleration(particles)
+
+        # update velocity using acceleration vector
+        self.vx += self.ax * dt
+        self.vy += self.ay * dt
+        self.vz += self.az * dt
+
+        # update position using velocity vector
+        self.x += self.vx * dt
+        self.y += self.vy * dt
+        self.z += self.vz * dt
+
+
+
+    def acceleration(self, particles):
+        G = 0.0000000000066;
+
+        self.ax = 0;
+        for p in particles:
+            if(p != self):
[…]

### 2014-09-21 code/2body.py
Fixes initial values, however if two object collide the velocity takes extreme values

@@ -23,17 +23,18 @@ def init_opengl(width, height):
 class Particle:
     def __init__(self):
         # generate random start position
-        self.x = random.random() * 2 - 5
-        self.y = random.random() * 2 - 5
-        self.z = random.random() * -10
+        mag = 3
+        self.x = ((random.randrange(-100,100))/100)*mag
+        self.y = ((random.randrange(-100,100))/100)*mag
+        self.z = ((random.randrange(-100,100))/100)*mag
 
         # generate random start velocity
-        magnitude = 0.1
-        self.vx = (random.random() - 0.5)*magnitude
-        self.vy = (random.random() - 0.5)*magnitude
-        self.vz = (random.random() - 0.5)*magnitude
+        magnitude = 0.01
+        self.vx = ((random.randrange(-100,100))/100)*magnitude
+        self.vy = ((random.randrange(-100,100))/100)*magnitude
+        self.vz = ((random.randrange(-100,100))/100)*magnitude
 
-        self.mass = (random.random())*1000000;
+        self.mass = (random.random())*1000000000;
         self.ax = 0;
         self.ay = 0;
         self.az = 0;
@@ -44,7 +45,7 @@ class Particle:
         #glColor3f(1,1,1)
         #glColor3f(math.sqrt(math.pow(self.vx,2)), math.sqrt(math.pow(self.vy,2)), math.sqrt(math.pow(self.vz,2)))
         #glVertex3f(self.x, self.y, self.z)
-        glPointSize((self.mass/1000000)*10)
+        glPointSize((self.mass/1000000000)*10)
         glBegin(GL_POINTS)
         glColor3f(1,1,1)
         glVertex3f(self.x, self.y, -10);
@@ -64,6 +65,8 @@ class Particle:
         self.vy += self.ay * dt
         self.vz += self.az * dt
 
+        print(self.az)
+
         # update position using velocity vector
         self.x += self.vx * dt
         self.y += self.vy * dt
@@ -77,21 +80,21 @@ class Particle:
         self.ax = 0;
         for p in particles:
             if(p != self):
-                self.ax = self.ax + p.mass*(p.x - self.x)/math.pow(math.fabs(p.x - self.x),3)
+                self.ax = self.ax + p.mass*(p.x - self.x)/max(0.00000001, math.pow(math.fabs(p.x - self.x),3))
 
         self.ax = G*self.ax;
 
         self.ay = 0;
         for p in particles:
             if(p != self):
[…]

### 2014-09-23 code/2body.py
fixes simulation

@@ -23,16 +23,16 @@ def init_opengl(width, height):
 class Particle:
     def __init__(self):
         # generate random start position
-        mag = 3
-        self.x = ((random.randrange(-100,100))/100)*mag
-        self.y = ((random.randrange(-100,100))/100)*mag
-        self.z = ((random.randrange(-100,100))/100)*mag
+        mag = 10
+        self.x = ((random.randrange(-100,100))/100.0)*mag
+        self.y = ((random.randrange(-100,100))/100.0)*mag
+        self.z = ((random.randrange(-100,100))/100.0)*mag
 
         # generate random start velocity
         magnitude = 0.01
-        self.vx = ((random.randrange(-100,100))/100)*magnitude
-        self.vy = ((random.randrange(-100,100))/100)*magnitude
-        self.vz = ((random.randrange(-100,100))/100)*magnitude
+        self.vx = ((random.randrange(-100,100))/100.0)*magnitude
+        self.vy = ((random.randrange(-100,100))/100.0)*magnitude
+        self.vz = ((random.randrange(-100,100))/100.0)*magnitude
 
         self.mass = (random.random())*1000000000;
         self.ax = 0;
@@ -56,7 +56,6 @@ class Particle:
 
     def update(self, particles):
         dt = 0.1
-        gravity_strength = 0.5
 
         self.acceleration(particles)
 
@@ -78,25 +77,20 @@ class Particle:
         G = 0.0000000000066;
 
         self.ax = 0;
-        for p in particles:
-            if(p != self):
-                self.ax = self.ax + p.mass*(p.x - self.x)/max(0.00000001, math.pow(math.fabs(p.x - self.x),3))
-
-        self.ax = G*self.ax;
-
         self.ay = 0;
-        for p in particles:
-            if(p != self):
-                self.ay = self.ay + p.mass*(p.y - self.y)/max(0.00000001,math.pow(math.fabs(p.y - self.y),3))
-
-        self.ay = G*self.ay;
-        
         self.az = 0;
         for p in particles:
             if(p != self):
-                self.az = self.az + p.mass*(p.z - self.z)/max(0.00000001,math.pow(math.fabs(p.z - self.z),3))
+                r = math.sqrt((p.x - self.x)**2 + (p.y - self.y)**2 + (p.z - self.z)**2)
+                self.ax = self.ax + p.mass*((p.x - self.x)/max(0.00000001, math.pow(r,3)))
[…]

### 2014-09-23 article/article.tex
Updates readme file
article has description for calculating force and acceleration

@@
 \begin{equation}
+    \label{eq:newtongravity}
  F = G\frac{m_1m_2}{r^2}
@@
 The center of mass can be calculated in reference of one of our bodies.
+(bary center)
 \begin{equation}
@@
 \end{equation}
+force times direction (unit vector) see albana notes on slide
+Calculate $a_i$ for each body i.
 
+If we want to know the force of body $p$ on body $c$ we can use Newton's law of gravitation, see equation~\ref{eq:newtongravity}.
+We can fill in the masses of bodies $p,c$ using the fact that $F_c = m_c a_c$ we obtain
+\begin{equation}
+    \begin{split}
+     a_c m_c = G\frac{m_c m_p}{r^2} \\
+     a_c = G\frac{m_p}{r^2} \\
+    \end{split}
+\end{equation}
+Now that we have derived the acceleration in terms of the mass of $p$ and the distance between the bodies $r$, we can use this calculate the acceleration of body $c$.
+The change of position i.e.
+distance of body $c$ at time $\delta t$ is given by
+\begin{equation}
+    \frac{d^2}{dt^2} r_c = G \frac{m_p}{r^2}
+\end{equation}
+So the new position can be derived by taking the position at time $t=0$ plus the distance traveled during $\delta t$.
+This can be obtained by taking the double integral of the acceleration.
+This can be solved analytically when only considering two bodies.
+If you consider $n \geq 3$ bodies it becomes unsolvable analytically.
 \subsection{Discrete Approximation}
 Approximation of Newton's law, Kepler orbit. Possible accuracy problems, e.g. sensitivity analysis.
-
+\\\\
+Possible checks for simulation correctness: law of momentum conservation, law of energy conservation, unintended orbit change
 \subsection{Algorithms}

### 2014-09-26 code/cppsimulator/main.cpp
basic outline for C++ app with SDL2 and OpenGL

@@ -1,0 +1,93 @@
+#include <iostream>
+#include <OpenGL/gl.h>
+#include <OpenGL/glu.h>
+#include "SDL.h"
+
+SDL_Window* initSDLAndOpenGL(int width, int height, bool vsync, bool fullscreen)
+{
+    SDL_Init(SDL_INIT_VIDEO);
+    SDL_GL_SetAttribute( SDL_GL_RED_SIZE, 8 );
+    SDL_GL_SetAttribute( SDL_GL_GREEN_SIZE, 8 );
+    SDL_GL_SetAttribute( SDL_GL_BLUE_SIZE, 8 );
+    SDL_GL_SetAttribute( SDL_GL_DEPTH_SIZE, 32 );
+    SDL_GL_SetAttribute( SDL_GL_DOUBLEBUFFER, 1 );
+    SDL_GL_SetAttribute( SDL_GL_MULTISAMPLEBUFFERS , 1 );
+    SDL_GL_SetAttribute( SDL_GL_MULTISAMPLESAMPLES , 4 );
+    SDL_Window* screen;
+    if (fullscreen)
+    {
+        screen = SDL_CreateWindow("AHN2 Viewer",
+                SDL_WINDOWPOS_UNDEFINED,
+                SDL_WINDOWPOS_UNDEFINED,
+                width, height,
+                SDL_WINDOW_OPENGL | SDL_WINDOW_FULLSCREEN_DESKTOP);
+    }
+    else
+    {
+        screen = SDL_CreateWindow("AHN2 Viewer",
+                SDL_WINDOWPOS_UNDEFINED,
+                SDL_WINDOWPOS_UNDEFINED,
+                width, height,
+                SDL_WINDOW_OPENGL);
+    }
+
+    SDL_GLContext glcontext = SDL_GL_CreateContext(screen);
+
+    if (SDL_GL_SetSwapInterval(vsync? 1 : 0) == -1) {
+        std::cout << "could not set desired vsync mode: " << vsync << 
+            std::endl;
+    }
+    glClearColor(0.1, 0.1, 0.15, 0);
+
+    glClearDepth(1);
+    glDepthFunc(GL_LESS);
+
+    glViewport(0, 0, width, height);
+
+    glMatrixMode(GL_PROJECTION);
+    glLoadIdentity();
+    gluPerspective(90, float(width)/height, 0.1, 100.0);
+    glMatrixMode(GL_MODELVIEW);
+    glLoadIdentity();
[…]

### 2014-09-26 code/cppsimulator/main.cpp
simple particle rendering, no gravitational interaction

@@ -1,7 +1,9 @@
 #include <iostream>
+#include <vector>
 #include <OpenGL/gl.h>
 #include <OpenGL/glu.h>
 #include "SDL.h"
+#include "particle.h"
 
 SDL_Window* initSDLAndOpenGL(int width, int height, bool vsync, bool fullscreen)
 {
@@ -39,8 +41,8 @@ SDL_Window* initSDLAndOpenGL(int width, int height, bool vsync, bool fullscreen)
     }
     glClearColor(0.1, 0.1, 0.15, 0);
 
-    glClearDepth(1);
-    glDepthFunc(GL_LESS);
+    //glClearDepth(1);
+    //glDepthFunc(GL_LESS);
 
     glViewport(0, 0, width, height);
 
@@ -49,6 +51,7 @@ SDL_Window* initSDLAndOpenGL(int width, int height, bool vsync, bool fullscreen)
     gluPerspective(90, float(width)/height, 0.1, 100.0);
     glMatrixMode(GL_MODELVIEW);
     glLoadIdentity();
+    glPointSize(2);
 
 
     GLenum error = glGetError();
@@ -76,11 +79,37 @@ int main(int argc, char* argv[])
 
     SDL_Window* surface = initSDLAndOpenGL(screenWidth, screenHeight, vsync,
             fullscreen);
+
+    std::vector<Particle> particles;
+    for (int i = 0; i < 2000; i++)
+    {
+        particles.push_back(Particle());
+    }
+
+
     bool running = true;
+    float timestep = 0.1;
+    float move = 0;
     while (running)
     {
+        //run simulation
+        for(auto& particle: particles)
+        {
+            particle.update(timestep);
+        }
+
+        //render universe
         glClear(GL_COLOR_BUFFER_BIT | GL_DEPTH_BUFFER_BIT);
+        glLoadIdentity();
+        glTranslatef(0, 0, -30);
+        for(auto& particle: particles)
+        {
+            particle.render();
+        }
         SDL_GL_SwapWindow(surface);
+
+
+        //check input
         SDL_Event event;    
         while (SDL_PollEvent(&event)) 
         {

### 2014-09-26 code/cppsimulator/particle.cpp

@@ -1,0 +1,25 @@
+#include "particle.h"
+
+#include <iostream>
+#include <OpenGL/gl.h>
+
+Particle::Particle()
+{
+    _position = Vec3D::createRandom(-10, 10);
+    _velocity = Vec3D::createRandom(-0.1, 0.1);
+}
+
+
+void Particle::update(float timestep)
+{
+    _position += _velocity * timestep;
+    _velocity += _acceleration * timestep;
+}
+
+void Particle::render()
+{
+    glBegin(GL_POINTS);
+    glColor3f(1,1,1);
+    glVertex3f(_position.x(), _position.y(), _position.z());
+    glEnd();
+}

### 2014-09-26 code/cppsimulator/particle.h

@@ -1,0 +1,18 @@
+#ifndef PARTICLE_H_
+#define PARTICLE_H_
+
+#include "vec.h"
+
+class Particle
+{
+    public:
+        Particle();
+        void update(float timestep);
+        void render();
+    private:
+        Vec3D _position;
+        Vec3D _velocity;
+        Vec3D _acceleration;
+};
+
+#endif // PARTICLE_H_

### 2014-09-26 code/cppsimulator/vec.cpp

@@ -1,0 +1,40 @@
+#include "vec.h"
+
+#include <cstdlib>
+
+Vec3D::Vec3D():
+   _x(0), _y(0), _z(0) 
+{
+}
+
+Vec3D::Vec3D(float x, float y, float z):
+   _x(x), _y(y), _z(z) 
+{
+}
+
+Vec3D Vec3D::createRandom(Vec3D min, Vec3D max)
+{
+    float x = ((max._x-min._x)*((float)rand()/RAND_MAX))+min._x;
+    float y = ((max._y-min._y)*((float)rand()/RAND_MAX))+min._y;
+    float z = ((max._z-min._z)*((float)rand()/RAND_MAX))+min._z;
+    return Vec3D(x, y, z);
+}
+
+Vec3D Vec3D::createRandom(float min, float max)
+{
+    return createRandom(Vec3D(min, min, min), Vec3D(max, max, max));
+}
+
+
+Vec3D Vec3D::operator*(float scalar)
+{
+    return Vec3D(_x*scalar, _y*scalar, _z*scalar);
+}
+
+Vec3D& Vec3D::operator+=(const Vec3D& vec)
+{
+    _x += vec._x;
+    _y += vec._y;
+    _z += vec._z;
+    return (*this);
+}

### 2014-09-26 code/cppsimulator/vec.h

@@ -1,0 +1,21 @@
+#ifndef VEC_H_
+#define VEC_H_
+
+class Vec3D
+{
+    public:
+        Vec3D(float x, float y, float z);
+        Vec3D();
+        static Vec3D createRandom(Vec3D min, Vec3D max);
+        static Vec3D createRandom(float min, float max);
+        Vec3D operator*(float scalar);
+
+        Vec3D& operator+=(const Vec3D& vec);
+        float x() {return _x;}
+        float y() {return _y;}
+        float z() {return _z;}
+    private:
+        float _x, _y, _z;
+};
+
+#endif // VEC_H_

### 2014-09-26 code/cppsimulator/main.cpp
body gravity interaction, all bodies have same mass

@@ -89,10 +89,13 @@ int main(int argc, char* argv[])
 
     bool running = true;
     float timestep = 0.1;
-    float move = 0;
     while (running)
     {
         //run simulation
+        for(auto& particle: particles)
+        {
+            particle.calculateAcceleration(particles, timestep);
+        }
         for(auto& particle: particles)
         {
             particle.update(timestep);

### 2014-09-26 code/cppsimulator/particle.cpp

@@ -9,6 +9,22 @@ Particle::Particle()
     _velocity = Vec3D::createRandom(-0.1, 0.1);
 }
 
+void Particle::calculateAcceleration(const std::vector<Particle>& particles, float timestep)
+{
+    //set acceleration to zero vector
+    _acceleration = Vec3D();
+    const float G = 0.01;
+
+    for (const auto& particle: particles)
+    {
+        if (&particle == this)
+            continue;
+
+        Vec3D positionDelta = (particle._position - _position);
+        float l = positionDelta.length();
+        _acceleration +=  (positionDelta * timestep * G) / (l*l*l);
+    }
+}
 
 void Particle::update(float timestep)
 {

### 2014-09-26 code/cppsimulator/vec.cpp

@@ -1,6 +1,8 @@
 #include "vec.h"
 
 #include <cstdlib>
+#include <cmath>
+#include <iostream>
 
 Vec3D::Vec3D():
    _x(0), _y(0), _z(0) 
@@ -20,6 +22,27 @@ Vec3D Vec3D::createRandom(Vec3D min, Vec3D max)
     return Vec3D(x, y, z);
 }
 
+Vec3D Vec3D::operator+(const Vec3D& vec) const
+{
+    float x = _x + vec._x;
+    float y = _y + vec._y;
+    float z = _z + vec._z;
+    return Vec3D(x, y, z);
+}
+
+Vec3D Vec3D::operator-(const Vec3D& vec) const
+{
+    float x = _x - vec._x;
+    float y = _y - vec._y;
+    float z = _z - vec._z;
+    return Vec3D(x, y, z);
+}
+
+float Vec3D::length()
+{
+    return sqrt(_x*_x + _y*_y + _z*_z);
+}
+
 Vec3D Vec3D::createRandom(float min, float max)
 {
     return createRandom(Vec3D(min, min, min), Vec3D(max, max, max));
@@ -29,6 +52,11 @@ Vec3D Vec3D::createRandom(float min, float max)
 Vec3D Vec3D::operator*(float scalar)
 {
     return Vec3D(_x*scalar, _y*scalar, _z*scalar);
+}
+
+Vec3D Vec3D::operator/(float scalar)
+{
+    return Vec3D(_x/scalar, _y/scalar, _z/scalar);
 }
 
 Vec3D& Vec3D::operator+=(const Vec3D& vec)

### 2014-09-26 code/cppsimulator/main.cpp
add mouse rotation

@@ -89,6 +89,9 @@ int main(int argc, char* argv[])
 
     bool running = true;
     float timestep = 0.1;
+
+    float yRotate = 0;
+    float xRotate = 0;
     while (running)
     {
         //run simulation
@@ -105,6 +108,8 @@ int main(int argc, char* argv[])
         glClear(GL_COLOR_BUFFER_BIT | GL_DEPTH_BUFFER_BIT);
         glLoadIdentity();
         glTranslatef(0, 0, -30);
+        glRotatef(yRotate, 0,1,0);
+        glRotatef(xRotate, 1,0,0);
         for(auto& particle: particles)
         {
             particle.render();
@@ -119,6 +124,16 @@ int main(int argc, char* argv[])
             if (event.type == SDL_QUIT)
             {
                 running = false;
+            }
+            else if (event.type == SDL_MOUSEMOTION)
+            {
+                if (event.motion.state & SDL_BUTTON(1))
+                {
+                    yRotate += event.motion.xrel * 0.5;
+                    xRotate += event.motion.yrel * 0.5;
+                    if (xRotate > 90) xRotate = 90;
+                    if (xRotate < -90) xRotate = -90;
+                }
             }
         }
     }

### 2014-09-27 code/cppsimulator/main.cpp
fix rotation order, allow zooming using scrollwheel

@@ -92,6 +92,7 @@ int main(int argc, char* argv[])
 
     float yRotate = 0;
     float xRotate = 0;
+    float distance = -30;
     while (running)
     {
         //run simulation
@@ -107,9 +108,9 @@ int main(int argc, char* argv[])
         //render universe
         glClear(GL_COLOR_BUFFER_BIT | GL_DEPTH_BUFFER_BIT);
         glLoadIdentity();
-        glTranslatef(0, 0, -30);
-        glRotatef(yRotate, 0,1,0);
+        glTranslatef(0, 0, distance);
         glRotatef(xRotate, 1,0,0);
+        glRotatef(yRotate, 0,1,0);
         for(auto& particle: particles)
         {
             particle.render();
@@ -134,6 +135,10 @@ int main(int argc, char* argv[])
                     if (xRotate > 90) xRotate = 90;
                     if (xRotate < -90) xRotate = -90;
                 }
+            }
+            else if (event.type == SDL_MOUSEWHEEL)
+            {
+                distance += 0.3* event.wheel.y;
             }
         }
     }

### 2014-09-27 code/cppsimulator/main.cpp
fix window title

@@ -18,7 +18,7 @@ SDL_Window* initSDLAndOpenGL(int width, int height, bool vsync, bool fullscreen)
     SDL_Window* screen;
     if (fullscreen)
     {
-        screen = SDL_CreateWindow("AHN2 Viewer",
+        screen = SDL_CreateWindow("N-Body simulator",
                 SDL_WINDOWPOS_UNDEFINED,
                 SDL_WINDOWPOS_UNDEFINED,
                 width, height,
@@ -26,7 +26,7 @@ SDL_Window* initSDLAndOpenGL(int width, int height, bool vsync, bool fullscreen)
     }
     else
     {
-        screen = SDL_CreateWindow("AHN2 Viewer",
+        screen = SDL_CreateWindow("N-Body simulator",
                 SDL_WINDOWPOS_UNDEFINED,
                 SDL_WINDOWPOS_UNDEFINED,
                 width, height,

### 2014-09-28 article/article.tex
Adds text

@@
 
-\section{Introduction (phenomena)}
-A world famous problem is the two body problem, most commonly known as the sun - earth problem.
+\section{Introduction}
+Commonly used technology in everyday life is supported by satellites, e.g. GPS in car navigation, Internet and pictures from Google Maps. 
+And with increasingly more satellites orbiting the earth it is important that we know that a new satellite will not crash into another.
+Or, in more general terms we need to know how objects orbiting other objects work such that we can launch more satellites.
 
-\section{Literature Study}
-background etc
-\subsection{Newton's law}
+Although, it is not a new problem, because we have been trying to analyze how planets orbit each other in our own solar system for quite while, there is still room for improvement.
+The analytical solution for orbiting planets are only purposed for 2-body system, while in reality we have to cope with N-body systems.
+Approximation methods can be used only calculating is intractable by hand.
+Hence, the use of computers is promising for solving N-body systems.
+However, using computers and approximation methods comes with different types of problems, i.e.
+how accurate is my approximation method as well as the physical limitations of computers in terms of floating point operations.
+
+This research is aims to provide insight in the approximation methods used for 2-body systems, i.e. how accurate is the simulation. 
+In section~\ref{sec:literature} the background of the analytical models used are explained and section~\ref{sec:methodology} introduces the algorithms as well a methods for benchmarking our approximation methods.
+The results are shown in section~\ref{sec:results} and section~\ref{sec:conclusion} our conclusion is discussed.
+
+\section{Background}
+\label{sec:literature}
+In pursuance of calculating the trajectory of a body it is necessary to know which forces interact with our body.
+Newton's seconds law, i.e.
+the rate of change is equal to the net force on a object, is used to describe the force on our object.
+And Newton's third law, i.e.
+action = - reaction, is used.
+
+\subsection{Newton's laws}
+Newton's second law describes that the force on an object is equal to the change of momentum in time.
+Because we only consider constant mass systems this is equal to the mass times the acceleration of an object.
+Thus, if the object is accelerating there is an force on our object present.
+Knowing that the only possible object acting on our object are the other planets in the system we can modify the second law to derive the law of universal gravitation.
+Let the following equation be the force $F_p$ on object $p$ by object $c$ with mass $m_p,m_c$ respectively.
 \begin{equation}
     \label{eq:newtongravity}
- F = G\frac{m_1m_2}{r^2}
+ F_p = G\frac{m_pm_c}{r^2}
 \end{equation}
-With G as a gravitational constant and F as the force in Newton and $m_1,m_2$ the mass of the two systems in kg and $r$ is the distance between the center of the two bodies (see figure xx). 
-Keplers problem? 
-\subsection{Kepler}
-Kepler problem, special case the Kepler orbit problem
+With $r$ as the euclidean distance between object $p,c$. Combining Newton's second law with the law of universal gravitation gives the acceleration $a_p$ of object $p$
+\begin{equation}
+    \label{eq:newtongravity}
+	a_p = G\frac{m_pm_c}{ m_pr^2}
+\end{equation}.
+The third law states that if there is a force $F_1$ then there also exists a force $F_2$ such that $F_2 = - F_1$.
+
+In the next section a simple 2-body system is introduced for thorough understanding the approximation methods used in later sections.
 
 \section{Methodology}
-Assume that we have an initial velocity vector $\overrightarrow{v_t}$ for the body t with speed in both x and y direction.
-\begin{equation}
-	\overrightarrow{v_t} = \begin{bmatrix}
-								x_t \\
-								y_t
-							\end{bmatrix}
-\end{equation}
-For example we have a two body system with $p,c$ with an initial velocity vector, respectively, then
-\begin{equation}
-	v_p=\begin{bmatrix} v_x^p \\ v_y^p \end{bmatrix}, v_c=\begin{bmatrix} v_x^c \\ v_y^c \end{bmatrix}
-\end{equation}
+\label{sec:methodology}
+To understand the consequences of our approximation methods we first need to understand the analytical solution of a 2-body system.
+Next, we assume a simple two body system with an analytical solution and we will apply Newton's law to obtain the trajectory of our objects.
+Remark that this is special simple case designed to provided better understanding of orbiting objects, this is required to understand the effects of discrete approximation methods in our simulations.
+\\
+\indent Assume we have a 2 body system called Pluto-Charon system. Figure~\ref{fig:plutocharon} illustrates this system at a fixed time point $t$
+\begin{figure*}
+	\label{fig:plutocharon}
+\end{figure*}
+with both $x,y$ coordinates in our 2D space.
+
+% Assume that we have an initial velocity vector $\overrightarrow{v_t}$ for the body t with speed in both x and y direction.
+% \begin{equation}
+% 	\overrightarrow{v_t} = \begin{bmatrix}
+% 								x_t \\
+% 								y_t
+% 							\end{bmatrix}
+% \end{equation}
+% For example we have a two body system with $p,c$ with an initial velocity vector, respectively, then
+% \begin{equation}
+% 	v_p=\begin{bmatrix} v_x^p \\ v_y^p \end{bmatrix}, v_c=\begin{bmatrix} v_x^c \\ v_y^c \end{bmatrix}
+% \end{equation}
 We assume that the whole system has a constant velocity, in our case we assume zero velocity, hence the next equation must be equal to zero
@@
 \subsection{Discrete Approximation}
-Approximation of Newton's law, Kepler orbit. Possible accuracy problems, e.g. sensitivity analysis.
-\\\\
+The N-body system for more than 2 bodies can not be solved analytically, hence we must use approximation methods to calculate the solution.
+We want to obtain the new positions of our bodies at time $t+1$ by using Newton's second law. We can approximate
+
+\subsubsection{Taylor expansion}
+We want to approximate $a,v,x$ with Taylor expansions. 
+Because we want to approximate the calculation 
+Bounded error of Taylor approximation
+
+\subsubsection{Sensitivity analysis}
+because we use approximations we have sensitive systems etc....
+perturbations in input variables and output variables. i.e. positions of other objects are incorrect.
+Delta step influences sensitivity?....
+
+\subsubsection{Verify our approximation}
+Constraints check after each simulation round
 Possible checks for simulation correctness: law of momentum conservation, law of energy conservation, unintended orbit change
+
 \subsection{Algorithms}
+Algorithms in pseudo code:
+Taylor algorithm
+etc..
 
 \subsection{Implementation}
+attachment
 \section{Experimental Results}
+\label{sec:results}
+results
+
 \section{Conclusions}
+\label{sec:conclusion}
 

### 2014-09-28 code/2body.py

@@ -29,7 +29,7 @@ class Particle:
         self.z = ((random.randrange(-100,100))/100.0)*mag
 
         # generate random start velocity
-        magnitude = 0.01
+        magnitude = 0.5
         self.vx = ((random.randrange(-100,100))/100.0)*magnitude
         self.vy = ((random.randrange(-100,100))/100.0)*magnitude
         self.vz = ((random.randrange(-100,100))/100.0)*magnitude
@@ -55,7 +55,7 @@ class Particle:
         return self.mass;
 
     def update(self, particles):
-        dt = 0.1
+        dt = 0.01
 
         self.acceleration(particles)
 
@@ -64,7 +64,6 @@ class Particle:
         self.vy += self.ay * dt
         self.vz += self.az * dt
 
-        print(self.az)
 
         # update position using velocity vector
         self.x += self.vx * dt

### 2014-09-28 code/cppsimulator/main.cpp

@@ -81,7 +81,7 @@ int main(int argc, char* argv[])
             fullscreen);
 
     std::vector<Particle> particles;
-    for (int i = 0; i < 2000; i++)
+    for (int i = 0; i < 200000; i++)
     {
         particles.push_back(Particle());
     }

### 2014-09-30 code/cppsimulator/main.cpp
verification calculations to check c onservation of momentum and energy, visualisation of potential energy and particle paths

@@ -81,7 +81,7 @@ int main(int argc, char* argv[])
             fullscreen);
 
     std::vector<Particle> particles;
-    for (int i = 0; i < 200000; i++)
+    for (int i = 0; i < 2000; i++)
     {
         particles.push_back(Particle());
     }
@@ -95,6 +95,24 @@ int main(int argc, char* argv[])
     float distance = -30;
     while (running)
     {
+        //run model verification steps
+        Vec3D totalMomentum;
+        float totalKineticEnergy, totalPotentialEnergy;
+        for(auto& particle: particles)
+        {
+            totalMomentum = totalMomentum + particle.momentum();
+            totalKineticEnergy += particle.kineticEnergy();
+            totalPotentialEnergy += particle.potentialEnergy();
+        }
+        totalPotentialEnergy *= -0.5;
+
+
+        std::cout << "total momentum: (" << totalMomentum.length() << ") " ;
+        totalMomentum.print();
+        std::cout << "total kinetic energy: " << totalKineticEnergy << std::endl;
+        std::cout << "total potential energy: " << totalPotentialEnergy << std::endl;
+        std::cout << "total energy: " << totalPotentialEnergy + totalKineticEnergy  << std::endl;
+
         //run simulation
         for(auto& particle: particles)
         {

### 2014-09-30 code/cppsimulator/particle.cpp

@@ -7,12 +7,15 @@ Particle::Particle()
 {
     _position = Vec3D::createRandom(-10, 10);
     _velocity = Vec3D::createRandom(-0.1, 0.1);
+    _mass = 1;
 }
 
 void Particle::calculateAcceleration(const std::vector<Particle>& particles, float timestep)
 {
     //set acceleration to zero vector
     _acceleration = Vec3D();
+
+    _potentialEnergy = 0;
     const float G = 0.01;
 
     for (const auto& particle: particles)
@@ -22,7 +25,9 @@ void Particle::calculateAcceleration(const std::vector<Particle>& particles, flo
 
         Vec3D positionDelta = (particle._position - _position);
         float l = positionDelta.length();
+        _potentialEnergy += 1 / (l*l);
         _acceleration +=  (positionDelta * timestep * G) / (l*l*l);
+
     }
 }
 
@@ -30,12 +35,44 @@ void Particle::update(float timestep)
 {
     _position += _velocity * timestep;
     _velocity += _acceleration * timestep;
+
+    const int historySize = 20;
+    _positionHistory.push_back(_position);
+    if (_positionHistory.size() > historySize)
+    {
+        _positionHistory.pop_front();
+    }
 }
 
 void Particle::render()
 {
+    float redness = _potentialEnergy / 100;
+    if (redness > 1)
+        redness = 1;
+    glColor3f(1,1-redness,1-redness);
     glBegin(GL_POINTS);
-    glColor3f(1,1,1);
     glVertex3f(_position.x(), _position.y(), _position.z());
     glEnd();
+
+    glBegin(GL_LINE_STRIP);
+    int i = 0;
+    for (auto position: _positionHistory)
+    {
+        float c = float(i) / _positionHistory.size();
+        glColor3f(c,c,c);
+        i++;
+        glVertex3f(position.x(), position.y(), position.z());
+    }
+    glEnd();
+}
+
+Vec3D Particle::momentum()
+{
+    return _velocity * _mass;
+}
+
+float Particle::kineticEnergy()
+{
+    float v = _velocity.length();
+    return 0.5 * _mass * v * v;
 }

### 2014-09-30 code/cppsimulator/particle.h

@@ -2,6 +2,7 @@
 #define PARTICLE_H_
 
 #include <vector>
+#include <deque>
 #include "vec.h"
 
 class Particle
@@ -12,9 +13,15 @@ class Particle
         void update(float timestep);
         void render();
         Vec3D _position;
+        Vec3D momentum();
+        float kineticEnergy();
+        float potentialEnergy() {return _potentialEnergy;}
     private:
+        std::deque<Vec3D> _positionHistory;
+        float _potentialEnergy;
         Vec3D _velocity;
         Vec3D _acceleration;
+        float _mass;
 };
 
 #endif // PARTICLE_H_

### 2014-09-30 code/cppsimulator/vec.cpp

@@ -66,3 +66,8 @@ Vec3D& Vec3D::operator+=(const Vec3D& vec)
     _z += vec._z;
     return (*this);
 }
+
+void Vec3D::print()
+{
+    std::cout << _x << " " << _y << " " << _z << std::endl;
+}

### 2014-09-30 code/cppsimulator/vec.h

@@ -13,6 +13,7 @@ class Vec3D
         Vec3D operator+(const Vec3D& vec) const;
         Vec3D operator-(const Vec3D& vec) const;
         float length();
+        void print();
 
         Vec3D& operator+=(const Vec3D& vec);
         float x() const {return _x;}

### 2014-09-30 code/cppsimulator/particle.h
make sure to use mass in certain calculations

@@ -15,6 +15,7 @@ class Particle
         Vec3D _position;
         Vec3D momentum();
         float kineticEnergy();
+        float mass() const {return _mass;}
         float potentialEnergy() {return _potentialEnergy;}
     private:
         std::deque<Vec3D> _positionHistory;

### 2014-09-30 article/article.tex
Adds taylor expansion

@@
 The N-body system for more than 2 bodies can not be solved analytically, hence we must use approximation methods to calculate the solution.
-We want to obtain the new positions of our bodies at time $t+1$ by using Newton's second law. We can approximate
+We want to obtain the new positions of our bodies at time $t+1$ by using Newton's second law.
+We can derive the next position by adding the distance traveled in time $\Delta t$ to the old position at $x_t$, i.e.
+$x_{t+1} = x_t + v_t*\Delta t$ wit $v_t$ as the speed at time $t$.
+Nonetheless, we do not know speed $v_t$ but this can be obtained by $v_t = v_t + a_t \Delta t$ and $a_t$ is known for our body.
+The next section describes an even more accurate approach for $x_t$.
 
 \subsubsection{Taylor expansion}
-We want to approximate $a,v,x$ with Taylor expansions. 
-Because we want to approximate the calculation 
-Bounded error of Taylor approximation
+If we want to approximate the values of, for example, formula $f(x)$ at point $a$ we can use a Taylor expansion defined as
+\begin{equation}
+    \label{eq:taylor}
+    \sum^\infty_{n=0} \frac{f^{(n)}(a)}{n!}(x-a)^n
+\end{equation}
+when we sum to infinity the new function will be the same as the original function $f(x)$.
+In our case we want to approximate the formulas $x(t), v(t), a(t)$ at points near $t$ and we can use equation~\ref{eq:taylor} to describe $x(t)$ as with
+\begin{equation}
+    x(t) = \sum^\infty_{n=0} \frac{x^{(n)}(t)}{n!}(x-t)^n
+\end{equation}
+however, we are dealing with limited precision and will only use a few terms of the summation such that $x(t)$ becomes
+\begin{equation}
+    x(t) =  \frac{x^{(0)}(t)}{0!}(x-t)^0 + \frac{x^{(1)}(t)}{1!}(x-t)^1 +\frac{x^{(2)}(t)}{2!}(x-t)^2 +\frac{x^{(3)}(t)}{3!}(x-t)^3
+\end{equation}
+notice that $x^{(1)}(t)$ is the derivative of $x(t)$ which is equal to the velocity $v(t)$ and $x^{(2)}$ is equal to the acceleration $a(t)$.
+But we can also formulate a Taylor expansion for $v(t)$ and $a(t)$
+\begin{equation}
+    \begin{split}
+        v(t) =  \frac{v^{(0)}(t)}{0!}(x-t)^0 + \frac{v^{(1)}(t)}{1!}(x-t)^1 +\frac{v^{(2)}(t)}{2!}(x-t)^2\\
+        v(t) = v(0) + a(t) + j(t)
+    \end{split}
+\end{equation}
+\begin{equation}
+    \begin{split}
+        a(t) =  \frac{a^{(0)}(t)}{0!}(x-t)^0 + \frac{a^{(1)}(t)}{1!}(x-t)^1 \\
+        a(t) =  a(0) + j(t) 
+    \end{split}
+\end{equation}
+we can calculate $j(t)$ by taking the derivative of equation~\ref{eq:newtongravity} with respect to $a$. Thus obtaining more accurate results for our $x(t), v(t)$ and $a(t)$.
+Also, error of the approximation can be calculated as follows:
 
@@
 \subsubsection{Verify our approximation}
+\textbf{Momentum, angular momentum}\\
+fdsfs \\
+\textbf{Energy}\\
 Constraints check after each simulation round

### 2014-10-19 code/cppsimulator/integrator.cpp
leapfrog integrator

@@ -122,3 +122,55 @@ void Integrator::naiveIntegrator(std::vector<Particle>& particles, bool dynamic_
         }
     }
 }
+
+void Integrator::leapfrogIntegrator(std::vector<Particle>& particles, bool dynamic_time, float initial_timestep)
+{
+    //run simulation
+    float globalMin;
+    for(auto& particle: particles)
+    {
+        particle._prevAcceleration = particle._acceleration;
+        particle._acceleration = Vec3D();
+
+        particle._potentialEnergy = 0;
+        const float G = 0.01;
+
+        for (const auto& p: particles)
+        {
+            if (&p == &particle)
+                continue;
+
+            Vec3D positionDelta = (p._position - particle._position);
+            Vec3D velocityDelta = p._velocity - particle._velocity;
+            float l = positionDelta.length() + 0.1;
+            particle._potentialEnergy += (particle._mass * p.mass() * G) / (l*l);
+            Vec3D accelerationDelta = (positionDelta * timestep * G * particle.mass()) / (l*l*l);
+            particle._acceleration += accelerationDelta;
+
+            float minV = positionDelta.length() / velocityDelta.length();
+            float minA = sqrt(positionDelta.length() / accelerationDelta.length());
+
+            globalMin = minV < globalMin ? minV : globalMin;
+            globalMin = minA < globalMin ? minA : globalMin;
+    }
+
+    
+    }
+
+    timestep = (dynamic_time)?fmin(globalMin, GLOBAL_MINIMUM_TIMESTEP):initial_timestep;
+    timestep = (dynamic_time)?fmin(timestep, GLOBAL_MAXIMUM_TIMESTEP):initial_timestep;
+
+    for(auto& particle: particles)
+    {
+        //Taylor expansion around the current point
+        particle._position = particle._position +  particle._velocity * timestep + particle._acceleration * 0.5 * timestep * timestep;
+        particle._velocity = particle._velocity + (particle._acceleration + particle._prevAcceleration) * 0.5 * timestep;
+
+        const int historySize = 20;
+        particle._positionHistory.push_back(particle._position);
+        if (particle._positionHistory.size() > historySize)
+        {
+            particle._positionHistory.pop_front();
+        }
+    }
+}

### 2014-10-19 code/cppsimulator/integrator.h

@@ -10,6 +10,7 @@ class Integrator
     public:
         Integrator();
         void naiveIntegrator(std::vector<Particle>& particles, bool dynamic_timestep, float initial_timestep);
+        void leapfrogIntegrator(std::vector<Particle>& particles, bool dynamic_timestep, float initial_timestep);
         void jerkIntegrator(std::vector<Particle>& particles, bool dynamic_timestep, float initial_timestep);
     private:
         float timestep;

### 2014-10-19 code/cppsimulator/main.cpp

@@ -95,7 +95,7 @@ int main(int argc, char* argv[])
     //create log file
     std::ofstream outfile("log_file.txt");
 
-    enum IntegrationType {NAIVE, JERK};
+    enum IntegrationType {NAIVE, JERK, LEAPFROG};
     bool dynamicTime = false;
 
     IntegrationType integrationType = NAIVE;
@@ -106,6 +106,8 @@ int main(int argc, char* argv[])
             integrationType = NAIVE;
         else if ("jerk" == std::string(argv[1]))
             integrationType = JERK;
+        else if ("leapfrog" == std::string(argv[1]))
+            integrationType = LEAPFROG;
         if(argc > 2){
             if ("dynamic" == std::string(argv[2]))
                 dynamicTime = true;
@@ -129,7 +131,7 @@ int main(int argc, char* argv[])
 
     int loop_count = 0;
 
-    while (loop_count < 10000)
+    while (loop_count < 10000 && running)
     {
         //run model verification steps
         Vec3D totalMomentum;
@@ -169,6 +171,9 @@ int main(int argc, char* argv[])
         {
             case JERK:
                 integrator.jerkIntegrator(particles,dynamicTime, 0.01);
+                break;
+            case LEAPFROG:
+                integrator.leapfrogIntegrator(particles,dynamicTime, 0.01);
                 break;
             case NAIVE:
             default:

### 2014-10-19 code/cppsimulator/particle.h

@@ -21,6 +21,7 @@ class Particle
         float _potentialEnergy;
         Vec3D _velocity;
         Vec3D _acceleration;
+        Vec3D _prevAcceleration;
         float _mass;
         Vec3D _jerk;
 };

### 2014-10-19 code/cppsimulator/integrator.h
constant time simulatie ipv constant iteratie count

@@ -12,6 +12,7 @@ class Integrator
         void naiveIntegrator(std::vector<Particle>& particles, bool dynamic_timestep, float initial_timestep);
         void leapfrogIntegrator(std::vector<Particle>& particles, bool dynamic_timestep, float initial_timestep);
         void jerkIntegrator(std::vector<Particle>& particles, bool dynamic_timestep, float initial_timestep);
+        float getTimestep() {return timestep;}
     private:
         float timestep;
 };

### 2014-10-19 code/cppsimulator/main.cpp

@@ -129,9 +129,9 @@ int main(int argc, char* argv[])
     Integrator integrator;
 
 
-    int loop_count = 0;
+    float totalTime = 0;
 
-    while (loop_count < 10000 && running)
+    while (totalTime < 1000 && running)
     {
         //run model verification steps
         Vec3D totalMomentum;
@@ -165,7 +165,7 @@ int main(int argc, char* argv[])
             return -1;
         }
 
-        outfile << totalKineticEnergy << "," << totalPotentialEnergy << "," << relativeEnergyError << ";" << std::endl;
+        outfile << totalKineticEnergy << "," << totalPotentialEnergy << "," << relativeEnergyError << "," << totalTime <<  ";" << std::endl;
 
         switch (integrationType)
         {
@@ -219,7 +219,7 @@ int main(int argc, char* argv[])
             }
         }
 
-        loop_count++;
+        totalTime += integrator.getTimestep();
     }
     outfile.close();
 }

### 2014-10-20 code/cppsimulator/main.cpp
rendering flag, gnuplot code

@@ -97,22 +97,25 @@ int main(int argc, char* argv[])
 
     enum IntegrationType {NAIVE, JERK, LEAPFROG};
     bool dynamicTime = false;
+    bool norender = false;
 
     IntegrationType integrationType = NAIVE;
 
     if (argc > 1)
     {
-        if ("naive" == std::string(argv[1]))
+        for (int i = 1; i < argc; i++)
+        {
+        if ("naive" == std::string(argv[i]))
             integrationType = NAIVE;
-        else if ("jerk" == std::string(argv[1]))
+        else if ("jerk" == std::string(argv[i]))
             integrationType = JERK;
-        else if ("leapfrog" == std::string(argv[1]))
+        else if ("leapfrog" == std::string(argv[i]))
             integrationType = LEAPFROG;
-        if(argc > 2){
-            if ("dynamic" == std::string(argv[2]))
-                dynamicTime = true;
+        else if ("dynamic" == std::string(argv[i]))
+            dynamicTime = true;
+        else if ("norender" == std::string(argv[i]))
+            norender = true;
         }
-        
     }
 
   
@@ -131,7 +134,7 @@ int main(int argc, char* argv[])
 
     float totalTime = 0;
 
-    while (totalTime < 1000 && running)
+    while (totalTime < 100 && running)
     {
         //run model verification steps
         Vec3D totalMomentum;
@@ -145,7 +148,7 @@ int main(int argc, char* argv[])
         totalPotentialEnergy *= -0.5;
 
         float totalEnergy = totalPotentialEnergy + totalKineticEnergy;
-        if (initialRun)
+        if (initialRun || totalTime < 0.02)
         {
             initialTotalEnergy = totalEnergy;
             initialRun = false;
@@ -170,29 +173,32 @@ int main(int argc, char* argv[])
         switch (integrationType)
         {
             case JERK:
-                integrator.jerkIntegrator(particles,dynamicTime, 0.01);
+                integrator.jerkIntegrator(particles,dynamicTime, 0.1);
                 break;
             case LEAPFROG:
-                integrator.leapfrogIntegrator(particles,dynamicTime, 0.01);
+                integrator.leapfrogIntegrator(particles,dynamicTime, 0.1);
                 break;
             case NAIVE:
             default:
-                integrator.naiveIntegrator(particles, dynamicTime, 0.01);
[…]

### 2014-10-20 code/cppsimulator/runsim.sh

@@ -1,0 +1,12 @@
+#!/bin/bash          
+rm log_file.txt
+./sim leapfrog norender
+mv log_file.txt leapfrog.txt
+./sim jerk norender
+mv log_file.txt jerk.txt
+#./sim jerk dynamic norender
+#mv log_file.txt jerk_dynamic.txt
+./sim naive norender
+mv log_file.txt naive.txt
+#./sim naive dynamic norender
+#mv log_file.txt naive_dynamic.txt

### 2014-10-20 code/cppsimulator/integrator.cpp
kleine bugfix leapfrog (moet oude acceleration gebruiken)

@@ -163,7 +163,7 @@ void Integrator::leapfrogIntegrator(std::vector<Particle>& particles, bool dynam
     for(auto& particle: particles)
     {
         //Taylor expansion around the current point
-        particle._position = particle._position +  particle._velocity * timestep + particle._acceleration * 0.5 * timestep * timestep;
+        particle._position = particle._position +  particle._velocity * timestep + particle._prevAcceleration * 0.5 * timestep * timestep;
         particle._velocity = particle._velocity + (particle._acceleration + particle._prevAcceleration) * 0.5 * timestep;
 
         const int historySize = 20;

### 2014-10-20 code/cppsimulator/main.cpp
set start position, velocities, number of bodies and timestep using command line args

@@ -86,11 +86,6 @@ int main(int argc, char* argv[])
     SDL_Window* surface = initSDLAndOpenGL(screenWidth, screenHeight, vsync,
             fullscreen);
 
-    std::vector<Particle> particles;
-    for (int i = 0; i < 200; i++)
-    {
-        particles.push_back(Particle());
-    }
 
     //create log file
     std::ofstream outfile("log_file.txt");
@@ -100,28 +95,50 @@ int main(int argc, char* argv[])
     bool norender = false;
 
     IntegrationType integrationType = NAIVE;
+    float positionSize = 10;
+    float speedSize = 0.1;
+    float timestep = 0.01;
+    int bodies = 200;
 
     if (argc > 1)
     {
         for (int i = 1; i < argc; i++)
         {
-        if ("naive" == std::string(argv[i]))
-            integrationType = NAIVE;
-        else if ("jerk" == std::string(argv[i]))
-            integrationType = JERK;
-        else if ("leapfrog" == std::string(argv[i]))
-            integrationType = LEAPFROG;
-        else if ("dynamic" == std::string(argv[i]))
-            dynamicTime = true;
-        else if ("norender" == std::string(argv[i]))
-            norender = true;
+            std::string arg(argv[i]);
+
+            std::string prefix("position=");
+            if (!arg.compare(0, prefix.size(), prefix))
+                positionSize = atof(arg.substr(prefix.size()).c_str());
+            prefix = "speed=";
+            if (!arg.compare(0, prefix.size(), prefix))
+                speedSize = atof(arg.substr(prefix.size()).c_str());
+            prefix = "bodies=";
+            if (!arg.compare(0, prefix.size(), prefix))
+                bodies = atoi(arg.substr(prefix.size()).c_str());
+            prefix = "timestep=";
+            if (!arg.compare(0, prefix.size(), prefix))
+                timestep = atof(arg.substr(prefix.size()).c_str());
+            else if ("naive" == arg)
+                integrationType = NAIVE;
+            else if ("jerk" == arg)
+                integrationType = JERK;
+            else if ("leapfrog" == arg)
+                integrationType = LEAPFROG;
+            else if ("dynamic" == arg)
+                dynamicTime = true;
+            else if ("norender" == arg)
+                norender = true;
         }
     }
 
   
+    std::vector<Particle> particles;
+    for (int i = 0; i < 200; i++)
+    {
+        particles.push_back(Particle(positionSize, speedSize));
+    }
 
[…]

### 2014-10-20 code/cppsimulator/particle.cpp

@@ -3,14 +3,17 @@
 #include <iostream>
 #include <OpenGL/gl.h>
 
-Particle::Particle()
+Particle::Particle() : Particle(1, 0.01)
 {
-    _position = Vec3D::createRandom(-10, 10);
-    _velocity = Vec3D::createRandom(-1, 1);
+}
+
+Particle::Particle(float positionSize, float velocitySize)
+{
+    _position = Vec3D::createRandom(-positionSize, positionSize);
+    _velocity = Vec3D::createRandom(-velocitySize, velocitySize);
     _acceleration = Vec3D();
     _mass = 1;
 }
-
 
 
 

### 2014-10-20 code/cppsimulator/particle.h

@@ -9,6 +9,7 @@ class Particle
 {
     public:
         Particle();
+        Particle(float positionSize, float velocitySize);
         void render();
         Vec3D _position;
         Vec3D momentum();

### 2014-10-20 code/cppsimulator/runsim.sh
euler & leapfrog with different timesteps

@@ -1,12 +1,31 @@
 #!/bin/bash          
 rm log_file.txt
-./sim leapfrog norender
+#./sim leapfrog norender
 mv log_file.txt leapfrog.txt
 ./sim jerk norender
 mv log_file.txt jerk.txt
-#./sim jerk dynamic norender
-#mv log_file.txt jerk_dynamic.txt
+./sim jerk dynamic norender
+mv log_file.txt jerk_dynamic.txt
 ./sim naive norender
 mv log_file.txt naive.txt
-#./sim naive dynamic norender
-#mv log_file.txt naive_dynamic.txt
+./sim naive dynamic norender
+mv log_file.txt naive_dynamic.txt
+
+
+./sim naive norender timestep=0.001
+mv log_file.txt naive.0_001.txt
+./sim naive norender timestep=0.01
+mv log_file.txt naive.0_01.txt
+./sim naive norender timestep=0.1
+mv log_file.txt naive.0_1.txt
+./sim naive norender timestep=1
+mv log_file.txt naive.1.txt
+
+./sim leapfrog norender timestep=0.001
+mv log_file.txt leapfrog.0_001.txt
+./sim leapfrog norender timestep=0.01
+mv log_file.txt leapfrog.0_01.txt
+./sim leapfrog norender timestep=0.1
+mv log_file.txt leapfrog.0_1.txt
+./sim leapfrog norender timestep=1
+mv log_file.txt leapfrog.1.txt

### 2014-10-20 code/cppsimulator/main.cpp
sim can measure time, with different body counts

@@ -5,6 +5,7 @@
 #include "SDL.h"
 #include "particle.h"
 #include "integrator.h"
+#include "timer.h"
 #include <fstream>
 #include <cstdlib>
 
@@ -100,6 +101,9 @@ int main(int argc, char* argv[])
     float timestep = 0.01;
     int bodies = 200;
 
+    bool timeTrial = false;
+    int runs = 1;
+
     if (argc > 1)
     {
         for (int i = 1; i < argc; i++)
@@ -118,7 +122,13 @@ int main(int argc, char* argv[])
             prefix = "timestep=";
             if (!arg.compare(0, prefix.size(), prefix))
                 timestep = atof(arg.substr(prefix.size()).c_str());
-            else if ("naive" == arg)
+            prefix = "timetrial=";
+            if (!arg.compare(0, prefix.size(), prefix))
+            {
+                timeTrial = true;
+                runs = atoi(arg.substr(prefix.size()).c_str());
+            }
+            if ("naive" == arg)
                 integrationType = NAIVE;
             else if ("jerk" == arg)
                 integrationType = JERK;
@@ -131,118 +141,136 @@ int main(int argc, char* argv[])
         }
     }
 
-  
-    std::vector<Particle> particles;
-    for (int i = 0; i < 200; i++)
+    std::cout << "timestep:" <<  timestep << std::endl;
+
+    Timer timer;
+    timer.start();
+
+    for (int i = 0; i < runs; i++)
     {
-        particles.push_back(Particle(positionSize, speedSize));
-    }
-
-    bool running = true;
-
-    float yRotate = 0;
-    float xRotate = 0;
-    float distance = -30;
-    float initialTotalEnergy;
-    bool initialRun = true;
-
-    Integrator integrator;
-
-
-    float totalTime = 0;
-
-    while (totalTime < 100 && running)
-    {
-        //run model verification steps
-        Vec3D totalMomentum;
-        float totalKineticEnergy = 0, totalPotentialEnergy = 0;
-        for(auto& particle: particles)
+        std::vector<Particle> particles;
+        for (int i = 0; i < bodies; i++)
         {
-            totalMomentum = totalMomentum + particle.momentum();
-            totalKineticEnergy += particle.kineticEnergy();
-            totalPotentialEnergy += particle.potentialEnergy();
-        }
-        totalPotentialEnergy *= -0.5;
-
-        float totalEnergy = totalPotentialEnergy + totalKineticEnergy;
[…]

### 2014-10-20 code/cppsimulator/timer.h

@@ -1,0 +1,25 @@
+#ifndef TIMER_H_
+#define TIMER_H_
+#include <vector>
+
+const int CACHE_SIZE = 800;
+
+class Timer
+{
+  private:
+    int _ticks;
+    int _totalTicks;
+    int _tickCache[CACHE_SIZE];
+    int _currentCache;
+  public:
+    Timer();
+    void start();
+    double reset();
+    double stop();
+    double total();
+    std::vector<int> getTickHistory();
+
+    bool wait(int seconds);
+};
+
+#endif // TIMER_H_

### 2014-10-20 code/cppsimulator/runsim.sh
new test run, body count vs time

@@ -1,6 +1,6 @@
 #!/bin/bash          
 rm log_file.txt
-#./sim leapfrog norender
+./sim leapfrog norender
 mv log_file.txt leapfrog.txt
 ./sim jerk norender
 mv log_file.txt jerk.txt
@@ -29,3 +29,16 @@ mv log_file.txt leapfrog.0_01.txt
 mv log_file.txt leapfrog.0_1.txt
 ./sim leapfrog norender timestep=1
 mv log_file.txt leapfrog.1.txt
+
+./sim jerk norender timestep=0.001
+mv log_file.txt jerk.0_001.txt
+./sim jerk norender timestep=0.01
+mv log_file.txt jerk.0_01.txt
+./sim jerk norender timestep=0.1
+mv log_file.txt jerk.0_1.txt
+./sim jerk norender timestep=1
+mv log_file.txt jerk.1.txt
+
+
+./sim naive norender timetrial=11 timestep=0.1 bodies=2
+mv log_file.txt naive_time_bodies.txt

### 2014-10-20 code/cppsimulator/runsim.sh
add jerk to time vs bodies chart

@@ -1,44 +1,46 @@
 #!/bin/bash          
 rm log_file.txt
-./sim leapfrog norender
-mv log_file.txt leapfrog.txt
-./sim jerk norender
-mv log_file.txt jerk.txt
-./sim jerk dynamic norender
-mv log_file.txt jerk_dynamic.txt
-./sim naive norender
-mv log_file.txt naive.txt
-./sim naive dynamic norender
-mv log_file.txt naive_dynamic.txt
+#./sim leapfrog norender
+#mv log_file.txt leapfrog.txt
+#./sim jerk norender
+#mv log_file.txt jerk.txt
+#./sim jerk dynamic norender
+#mv log_file.txt jerk_dynamic.txt
+#./sim naive norender
+#mv log_file.txt naive.txt
+#./sim naive dynamic norender
+#mv log_file.txt naive_dynamic.txt
+#
+#
+#./sim naive norender timestep=0.001
+#mv log_file.txt naive.0_001.txt
+#./sim naive norender timestep=0.01
+#mv log_file.txt naive.0_01.txt
+#./sim naive norender timestep=0.1
+#mv log_file.txt naive.0_1.txt
+#./sim naive norender timestep=1
+#mv log_file.txt naive.1.txt
+#
+#./sim leapfrog norender timestep=0.001
+#mv log_file.txt leapfrog.0_001.txt
+#./sim leapfrog norender timestep=0.01
+#mv log_file.txt leapfrog.0_01.txt
+#./sim leapfrog norender timestep=0.1
+#mv log_file.txt leapfrog.0_1.txt
+#./sim leapfrog norender timestep=1
+#mv log_file.txt leapfrog.1.txt
+#
+#./sim jerk norender timestep=0.001
+#mv log_file.txt jerk.0_001.txt
+#./sim jerk norender timestep=0.01
+#mv log_file.txt jerk.0_01.txt
+#./sim jerk norender timestep=0.1
+#mv log_file.txt jerk.0_1.txt
+#./sim jerk norender timestep=1
+#mv log_file.txt jerk.1.txt
 
 
-./sim naive norender timestep=0.001
-mv log_file.txt naive.0_001.txt
-./sim naive norender timestep=0.01
-mv log_file.txt naive.0_01.txt
-./sim naive norender timestep=0.1
-mv log_file.txt naive.0_1.txt
-./sim naive norender timestep=1
-mv log_file.txt naive.1.txt
-
[…]

### 2014-10-20 code/cppsimulator/integrator.cpp
charts comparing dynamic vs constant timesteps

@@ -5,7 +5,7 @@
 #include <algorithm>
 
 #define GLOBAL_MINIMUM_TIMESTEP 0.1
-#define GLOBAL_MAXIMUM_TIMESTEP 0.0000001
+#define GLOBAL_MAXIMUM_TIMESTEP 0.001
 
 Integrator::Integrator() :
     timestep(0.1)

### 2014-10-20 code/cppsimulator/demo.sh
demo mode

@@ -1,0 +1,2 @@
+#!/bin/bash
+./sim demo=100 speed=0.15 time=40 timestep=0.1 bodies=300

### 2014-10-20 code/cppsimulator/main.cpp

@@ -100,9 +100,11 @@ int main(int argc, char* argv[])
     float speedSize = 0.1;
     float timestep = 0.01;
     int bodies = 200;
+    float maxTime = 100;
 
     bool timeTrial = false;
     int runs = 1;
+    bool demo = false;
 
     if (argc > 1)
     {
@@ -122,10 +124,19 @@ int main(int argc, char* argv[])
             prefix = "timestep=";
             if (!arg.compare(0, prefix.size(), prefix))
                 timestep = atof(arg.substr(prefix.size()).c_str());
+            prefix = "time=";
+            if (!arg.compare(0, prefix.size(), prefix))
+                maxTime = atof(arg.substr(prefix.size()).c_str());
             prefix = "timetrial=";
             if (!arg.compare(0, prefix.size(), prefix))
             {
                 timeTrial = true;
+                runs = atoi(arg.substr(prefix.size()).c_str());
+            }
+            prefix = "demo=";
+            if (!arg.compare(0, prefix.size(), prefix))
+            {
+                demo = true;
                 runs = atoi(arg.substr(prefix.size()).c_str());
             }
             if ("naive" == arg)
@@ -146,7 +157,9 @@ int main(int argc, char* argv[])
     Timer timer;
     timer.start();
 
-    for (int i = 0; i < runs; i++)
+    bool running = true;
+
+    for (int i = 0; i < runs && running; i++)
     {
         std::vector<Particle> particles;
         for (int i = 0; i < bodies; i++)
@@ -154,11 +167,12 @@ int main(int argc, char* argv[])
             particles.push_back(Particle(positionSize, speedSize));
         }
 
-        bool running = true;
 
         float yRotate = 0;
         float xRotate = 0;
         float distance = -30;
+        if (demo)
+            distance = -15;
         float initialTotalEnergy;
         bool initialRun = true;
 
@@ -168,7 +182,7 @@ int main(int argc, char* argv[])
         float totalTime = 0;
         timer.reset();
 
-        while (totalTime < 100 && running)
+        while (totalTime < maxTime && running)
         {
             //run model verification steps
             Vec3D totalMomentum;

### 2014-10-20 code/cppsimulator/particle.cpp

@@ -19,7 +19,7 @@ Particle::Particle(float positionSize, float velocitySize)
 
 void Particle::render()
 {
-    float redness = _potentialEnergy / 10;
+    float redness = _potentialEnergy * 20;
     if (redness > 1)
         redness = 1;
     glColor3f(1,1-redness,1-redness);

### 2014-10-24 article/article.tex
intro algorithm sectie

@@
 \subsection{Algorithms}
-Algorithms in pseudo code:
-Taylor algorithm
-etc..
+The n-body problem involves every body in a system interacting with every other body through gravity.
+This means that there are $\mathcal{O}(n^2)$ interactions that need to be calculated.
+Very broadly n-body simulators fall into two categories.
+The first category contains direct solvers, which simulate the gravitational interactions between all bodies directly, e.g.
+phiGRAPE \cite{Harfst2007357}.
+\\
+Alternatively, there are solvers that use a tree to represent the set of bodies in a hierarchical manner.
+These algorithms have each of the bodies interact with the tree instead of the full set of n-bodies.
+This can lead to a computational complexity of $\mathcal{O}(n \log(n))$.
+The Barnes-Hut algorithm is a pioneering example of this class of n-body simulation algorithms \cite{barnes1986hierarchical}.\\
+We have opted to go for a direct simulator, because the implementation of such algorithms tends to be less complex.
+The important part in these simulations is the determination of the force each of the bodies experiences.
+In direct solvers, this is done the same for each body to body interactions and the total force on a particular body is a simple summation of the forces it experiences from every other body.
+\\
+To determine the force between two bodies due to gravity is a second order differential equation needs to be solved. Thus numerical integrators are the key component of direct solvers.
+\subsubsection{Integrator Algorithms}
 
@@
 \label{sec:conclusion}
-
+\bibliographystyle{alpha}
+\bibliography{article}
 \end{document}

### 2014-10-24 article/article.tex
euler's method

@@
 \subsubsection{Integrator Algorithms}
-
+The integrators need to solve a second order differential equation.
+Based on calculating the force at a current point in time acceleration for a body can be determined.
+Based on this acceleration the position can be updated.
+The most straightforward way to do this is to apply Euler's method.
+Based on an acceleration and a stepsize $\Delta t$ it linearly approximates the speed at time $t + \Delta t$ from the speed at time $t$.
+Using the speed it then updates the position at time $t+\Delta t$ by linearly extrapolating from position at time $t$.
+\begin{equation}
+    \begin{split}
+    x(t+ \Delta t) = x(t) + v(t) \Delta t \\
+    v(t+ \Delta t) = v(t) + a(t) \Delta t \\
+    \end{split}
+\end{equation}
+This integrator is appealing for its simplicity, but because it is essentially a first order taylor expansion the expected error is proportional to $\Delta t^2$.
 \subsection{Implementation}

### 2014-10-24 article/article.tex
conservation of momentum

@@
 \subsubsection{Verify our approximation}
-\textbf{Momentum, angular momentum}\\
-fdsfs \\
+To verify the accuracy of the simulation we can apply some fundamental conservation laws from physics to our simulation.
+We know based on empirical data that these laws seem to be true for the natural, so if our simulation is a faithful model of nature these laws should also be true for our simulation.
+If they do not, the amount our simulation fails to conserve certain values that are supposed to be conserved can be used as a metric for the error in our simulation.\\\\
+\textbf{Conservation of Momentum}\\
+Momentum is defined as the mass of an object multiplied with its velocity vector. To get the momentum of a system of n bodies you can calculate the momentum of each body and sum them all together.
+\begin{equation}
+\vec{p}_{total} = \sum_i m_i \times \vec{v_i}
+\end{equation}
+As long as two interacting bodies experience equal but opposite forces the conservation of momentum tends to be pretty well preserved, even if those forces are otherwise wildly inaccurate.
+One the one hand this means that a fundamental law is enforced with any algorithm that ends up with the same force with a body to body interaction, just in opposite direction.
+This will be true for the n-body algorithms covered in this paper.
+On the other this means that the conservation of momentum is not a useful error metric to check the accuracy of an algorithm.
+Hence it is not used in this paper to compare the accuracy of the studied algorithms.
+The measurements have been done, but show no real difference, even when it is clear that a simulation is doing things that are unrealistic and inaccurate.
 \textbf{Energy}\\
@@
 \subsubsection{Integrator Algorithms}
+\textbf{Euler's Method}\\
 The integrators need to solve a second order differential equation.

### 2014-10-24 article/article.tex
conservation of energy

@@
 
-\subsubsection{Verify our approximation}
+\subsubsection{Model verification}
 To verify the accuracy of the simulation we can apply some fundamental conservation laws from physics to our simulation.
@@
 \begin{equation}
-\vec{p}_{total} = \sum_i m_i \times \vec{v_i}
+    \vec{p}_{total} = \sum_{i=1}^n m_i \times \vec{v_i}
 \end{equation}
-As long as two interacting bodies experience equal but opposite forces the conservation of momentum tends to be pretty well preserved, even if those forces are otherwise wildly inaccurate.
-One the one hand this means that a fundamental law is enforced with any algorithm that ends up with the same force with a body to body interaction, just in opposite direction.
+As long as two interacting bodies experience equal but opposite forces the conservation of momentum tends to be pretty well preserved, even if those forces are otherwise wildly inaccurate in a simulation.
+On the one hand this means that a fundamental law is enforced with any algorithm that ends up with the same force with a body to body interaction for both bodies, with only the direction reversed for one of them.
 This will be true for the n-body algorithms covered in this paper.
@@
 Hence it is not used in this paper to compare the accuracy of the studied algorithms.
-The measurements have been done, but show no real difference, even when it is clear that a simulation is doing things that are unrealistic and inaccurate.
-\textbf{Energy}\\
-Constraints check after each simulation round
-Possible checks for simulation correctness: law of momentum conservation, law of energy conservation, unintended orbit change
-
+The measurements have been done, but show no real difference, even when it is clear that a simulation is doing things that are otherwise unrealistic and inaccurate.\\\\
+\textbf{Conservation of Energy}\\
+Energy can not be created or destroyed, which in physics is expressed in the law of conservation of energy.
+Energy can be converted to a different type of energy, but the total amount of energy needs to be preserved.
+In a system of moving point masses that only gravitationally interact (i.e.
+a n-body system) the energy in the system is the sum of the kinetic energies and potential energies of each of the masses in the system.
+This sum should remain constant in a system that has no energy added or otherwise taken away from it.
+\\\\
+The kinetic energy of a body is the energy it has due to its motion. The total kinetic energy of a system is the kinetic energies of all its bodies summed together.
+\begin{equation}
+    K = \frac{1}{2} \sum_{i=1}^n m_i |\vec{v_i}|^2
+\end{equation}
+The potential energy of a body is the amount of energy it experiences due to its interaction with a force field, which would be with the gravitational field in a n-body simulation.
+The total potential energy of a system is the sum of the potential energy of all the bodies in the system.
+\begin{equation}
+    W = -\frac{1}{2} \sum_{i=1, i \neq j}^n G \frac{m_i m_j}{|r_i - r_j|}
+\end{equation}
+The total (mechanical) energy of system is the potential and kinetic energy of the system added together, $E = W+K$.
+This total energy should remain constant.
+In a simulation this can however go wrong.
+For instance a body may experience a certain strong potential energy.
+In reality it would only experience this energy for a very short time.
+But in a simulation with discrete timesteps, the timestep may be longer than the time the body is supposed to experience the potential energy for.
+Thus it will create extra kinetic energy without that energy being based on potential energy that would be there in reality when it interact continuously with bodies rather than at discrete timesteps.
+\\
+At the start of a simulation the initial energy in the system can be calculated.
+While running the simulation the energy can be calculated again.
+It should match the starting energy.
+The difference with the starting energy is the absolute energy error, and the difference divided by the current energy is the relative energy error.
+\begin{equation}
+    E_{error} = \frac{E_{current} - E_{initial}}{E_{current}}
+\end{equation}
+This relative energy error metric can be used to see how the size of the error progresses during a simulation with a particular algorithm and how the accuracy of one algorithm compares with an other algorithm.
 \subsection{Algorithms}

### 2014-10-25 article/article.tex
Fixes typos

@@
 
-This research is aims to provide insight in the approximation methods used for 2-body systems, i.e. how accurate is the simulation. 
+This research is aims to provide insight in the approximation methods used for N-body systems, i.e. how accurate is the simulation. 
 In section~\ref{sec:literature} the background of the analytical models used are explained and section~\ref{sec:methodology} introduces the algorithms as well a methods for benchmarking our approximation methods.
@@
 Because we only consider constant mass systems this is equal to the mass times the acceleration of an object.
-Thus, if the object is accelerating there is an force on our object present.
+Thus, if the object is accelerating there is a force on our object present.
 Knowing that the only possible object acting on our object are the other planets in the system we can modify the second law to derive the law of universal gravitation.
@@
 The product of those vectors gives us the directional vector of our body $X$.
-Due to the fact that the two bodies are the only objects in the system acting on eachother we can determine the radius and period of our movement around a (virtual) center of mass.
-We know for a fact that the two bodies must be the opposites of eachother such that the total momentum is equal to zero.
+Due to the fact that the two bodies are the only objects in the system acting on each other we can determine the radius and period of our movement around a (virtual) center of mass.
+We know for a fact that the two bodies must be the opposites of each other such that the total momentum is equal to zero.
 The center of mass can be calculated in reference of one of our bodies.
-(bary center)
+(Bary center)
 \begin{equation}
@@
 \end{equation}
-force times direction (unit vector) see albana notes on slide
-Calculate $a_i$ for each body i.
 
@@
 We can derive the next position by adding the distance traveled in time $\Delta t$ to the old position at $x_t$, i.e.
-$x_{t+1} = x_t + v_t*\Delta t$ wit $v_t$ as the speed at time $t$.
+$x_{t+1} = x_t + v_t*\Delta t$ with $v_t$ as the speed at time $t$.
 Nonetheless, we do not know speed $v_t$ but this can be obtained by $v_t = v_t + a_t \Delta t$ and $a_t$ is known for our body.

### 2014-10-25 article/article.tex
Adds graph and sections results

@@
 \subsubsection{Model verification}
+\label{verification}
 To verify the accuracy of the simulation we can apply some fundamental conservation laws from physics to our simulation.
@@
 \label{sec:results}
-results
+The results of each model implementation are compared using the metrics designed in section~\ref{verification}.
+Using these methods allows us to compare the performance of each model.
+For all the energy comparison methods we used a fixed total \textit{model evaluation time}, i.e.
+we simulate 100 model minutes.
+Depending on the time-steps used in each method, this might take more or less iterations to calculate, meaning that the total simulation time increases.
+In
+section~\ref{sec:res:euler} compares the relative error of the Euler method using different time-steps.
+Also, section~\ref{sec:res:taylor} compares the improvement of adding an extra term in the Taylor expansion.
+And, section~\ref{sec:res:leapfrog} shows the results using the Leapfrog integrator and all the models are compared in the last section~\ref{sec:res:all}.
+
+\subsection{Euler}
+\label{sec:res:euler}
+To verify our model, we have defined the energy metrics.
+In this experiment the relative energy metric as defined in section~\ref{XXX} is used.
+In total we have compared four different time-steps to evaluate the performance effect of an increasing/decreasing time-step.
+The time-steps used are $0.001, 0.01, 0.1, 1$ and figure~\ref{fig:euler} shows a plot of the relative energy error for the \textit{model simulation time}.
+\begin{figure}
+\label{fig:euler}
+    \includegraphics[width=0.9\textwidth]{euler_different_timesteps.png}
+    \caption{Relative energy error for the Euler integrator using different time-steps for a fixed total model simulation time.}
+\end{figure}
+
+\subsection{Taylor}
+\label{sec:res:taylor}
+
+\subsection{Leapfrog}
+\label{sec:res:leapfrog}
+
+\subsection{All models}
+\label{sec:res:all}
 

### 2014-10-25 article/article.tex
Adds jerk figure and tekst

@@
 \begin{figure}
-\label{fig:euler}
     \includegraphics[width=0.9\textwidth]{euler_different_timesteps.png}
     \caption{Relative energy error for the Euler integrator using different time-steps for a fixed total model simulation time.}
+    \label{fig:euler}
 \end{figure}
+As can be seen in figure~\ref{fig:euler} there is a spike around time 18 for time-step scale 1.
+This error has no apparent reason and in future time-steps it will decrease and stabilize around -1.
+Furthermore, it is interesting to see that with smaller time-step scales the total relative energy error increases faster.
+Meaning, that if we run the model for an infinite time the total error will explode and produce no meaning full results.
+This, while smaller time-step scales should increase the accuracy in terms of relative energy error.
+Therefore, we can conclude that the increase in accuracy does not outweighs the increase in iterations to complete the calculation.
 
@@
 \label{sec:res:taylor}
+Using the same energy metrics defined in section~\ref{XX} we can do the same simulation for the Taylor approximation.
+We expect that adding an extra term in the Taylor expansion would increase our accuracy.
+And in figure~\ref{fig:taylor} we can see that indeed slower increase in relative energy error than using bigger time-step scales.
+However, all the time-step scale converge to -1 after about 20\% of the total model simulation time.
+Also, we noticed that the 0.001 time-step scale converges faster to the -1 relative energy error than the 0.0001 time-step scale.
+Here, we can observe that decreasing the time-step by 10 does not increase the accuracy enough to compensate for the increase in number of iterations.
+The could be due to machine-precision, but requires further investigation to be sure.
+
+\begin{figure}
+    \includegraphics[width=0.9\textwidth]{jerk_different_timesteps_truncated.png}
+    \caption{Relative energy error for the Taylor integrator using different time-steps for a fixed total model simulation time.}
+    \label{fig:taylor}
+\end{figure}
 

### 2014-10-25 article/article.tex
Adds leapfrog results

@@
 \label{sec:res:leapfrog}
+Using the Leapfrog integrator, which is a slight modified Euler integrator, we should expect similar results.
+And as can be seen in figure~\ref{fig:leapfrog}.
+the relative energy error shows similar results as for the Euler approximation in figure~\ref{fig:euler}.
+However, the magnitude of the relative energy error is higher with the Leapfrog integrator at around time-step 18.
+This is probably due to the approximation of the in between time-steps which the Leapfrog integrator uses.
+Nevertheless, the results are relatively the same as the Euler approximation and does, at least in our simulation, lead to a decrease in relative energy error.
+
+\begin{figure}
+    \includegraphics[width=0.9\textwidth]{leapfrog_different_timesteps.png}
+    \caption{Relative energy error for the Leapfrog integrator using different time-steps for a fixed total model simulation time.}
+    \label{fig:leapfrog}
+\end{figure}
 
@@
 \label{sec:res:all}
+To compare all the models, i.e.
+Euler, Taylor and Leapfrog integrator, we setup up an experiment which uses the same time-step for all the models.
+The time-step is set at $0.01$ and figure~\ref{} shows the relative energy error for all models.
+Here we can clearly observe that the Taylor integrator gives the best results in terms of relative energy error.
+And as expected the Leapfrog and Euler integrator are approximately the same.
+However, the Taylor approximation comes with extra costs in computation time which will be discussed next.
+
+\begin{figure}
+    \includegraphics[width=0.9\textwidth]{constant_time.png}
+    \caption{Relative energy error for the Euler,Taylor and Leapfrog integrator using fixed time-steps for a fixed total model simulation time.}
+    \label{fig:leapfrog}
+\end{figure}
+
+\subsection{Complexity}
+\label{sec:res:complexity}
 

### 2014-10-25 article/article.tex
Adds conclusion

@@
 
+\subsection{Dynamic time-scales}
+Each of the models is also run with the dynamic time-scale, which calculates the global minimum time-step scale such that any two bodies can not ghost through one another.
+However, our experiments showed that after just a few iterations the dynamic time-scale converges to the minimum allowed time-step scale by our system.
+Hence, it is not useful to use the dynamic time-scales in the simulations.
+
 \subsection{All models}
@@
 And as expected the Leapfrog and Euler integrator are approximately the same.
-However, the Taylor approximation comes with extra costs in computation time which will be discussed next.
-
 \begin{figure}
@@
 \end{figure}
-
-\subsection{Complexity}
-\label{sec:res:complexity}
+However, the Taylor integrator comes with extra costs in computation time as can be seen in figure~\ref{fig:time}.
+The calculation time for calculating one extra Taylor term, i.e.
+the jerk, doubles the total computation time.
+Also, the figure shows that the increase in bodies has a quadratic behavior which is to be expected with $n^2$ complexity in the Euler integrator.
+\begin{figure}
+    \includegraphics[width=0.9\textwidth]{constant_time.png}
+    \caption{Computation time to finish calculating the total model simulation time for a given number of bodies.}
+    \label{fig:time}
+\end{figure}
 
@@
 \label{sec:conclusion}
+We have simulated different integrators and found a metric to compare the different models with each other.
+However, there is not one best model to select.
+Depending on the intentions of the simulation different models might be more applicable.
+For example, when speed has more priority than accuracy the Euler model might be more appropriate.
+Nevertheless, this research shows the complex considerations you have to take into account to build a simulation model.
+Also, that theoretical improvements do not always lead to actual improvements in a computer simulation.
+
+\section{Future work}
+
 \bibliographystyle{alpha}

### 2014-10-25 article/article.tex
Adds future work

@@
 \section{Future work}
+For future work we would like to recommend to try a time-step scale for each body, meaning that close encounters have smaller time-step scale than other bodies.
+This should lead to a smaller energy error while not slowing down the whole process.
+Also, many calculations are suitable for parallelization to speed up the computing process.
 

### 2014-10-25 article/article.tex
removes sensitivity analysis
jerk en leapfrog

--- 2014-10-25 article/article.tex
@@
     \begin{split}
-    x(t+ \Delta t) = x(t) + v(t) \Delta t \\
-    v(t+ \Delta t) = v(t) + a(t) \Delta t \\
+        x_{t+ 1} = x_t + v_t \Delta t \\
+        v_{t+ 1} = v_t + a_t \Delta t \\
     \end{split}
 \end{equation}
-This integrator is appealing for its simplicity, but because it is essentially a first order taylor expansion the expected error is proportional to $\Delta t^2$.
+This integrator is appealing for its simplicity, but because it is essentially a first order taylor expansion the expected error is proportional to $\Delta t^2$.\\\\
+\textbf{Jerk Integrator}\\
+It has been shown that through Newton's laws of gravity and mechanics you can calculate a body's acceleration based on the position of the other bodies.
+Acceleration is the second derivative of position, so it needs to be integrated twice to update the position of a body.
+\\\\
+It is however also possible to directly calculate the third derivative of the position, the so-called jerk.
+The jerk is also the first derivative of the acceleration and can thus be obtained by differentiating the acceleration once with respect to time.
+\begin{align}
+    \vec{j_i} = \frac{d}{dt} \vec{a_i} &= \frac{d}{dt} \left( G \sum_{k=1, k \neq i}^n \frac{m_k (\vec{r_k} - \vec{r_i})}{|\vec{r_k} - \vec{r_i}|^3} \right) \\
+                                       &= G \sum_{k=1, k \neq i} m_k \left( \frac{\vec{v_k} - \vec{v_i}}{|\vec{r_k} - \vec{r_i}|^3} - 3 \frac{(\vec{r_k} - \vec{r_i})\left((\vec{r_k} - \vec{r_i})\cdot (\vec{v_k} - \vec{v_i}) \right)}{|\vec{r_k} - \vec{r_i}|^5} \right)
+\end{align}
+Using the calculated jerk, the updated acceleration, velocity and position can be approximated.
+\begin{equation}
+    \begin{split}
+        x_{t+ 1} = x_t + v_t \Delta t + \frac{1}{2} a_t (\Delta t)^2 + \frac{1}{6} j_t (\Delta t)^3 \\
+        v_{t+ 1} = v_t a_t \Delta t + \frac{1}{2} j_t (\Delta t)^2 \\
+        a_{t+ 1} = a_t + j_t \Delta t \\
+    \end{split}
+\end{equation}
+\\\\
+\textbf{Leapfrog Integrator}\\
+The leapfrog integrator calculates the positions and velocities of the bodies in a simulated system at interleaved timesteps. 
+\begin{equation}
+    \begin{split}
+        x_{t+ 1} = x_t + v_{t+1/2} \Delta t \\
+        v_{t+ 1/2} = v_{t-1/2} + a_t \Delta t \\
+    \end{split}
+\end{equation}
+According to the literature (e.g. \cite{1995ApJ...443L..93H}) this can be rewritten in a form where all variables are evaluated at the same time.
+\begin{equation}
+    \begin{split}
+        x_{t+ 1} = x_t + v_t \Delta t  + \frac{1}{2} a_t (\Delta t)^2\\
+        v_{t+ 1} = v_t + \frac{a_t + a_{t+1}}{2} \Delta t \\
+    \end{split}
+\end{equation}
+One of the appealing characteristics of the leapfrog integrator is that it is time-symmetric.
+This means that if you let it run forward for a certain number of steps and then run it in reverse it should end up at the same initial starting state.
+This is supposed to improve its energy conservation properties.
 \subsection{Implementation}
@@
 \begin{figure}
+\label{fig:euler}
     \includegraphics[width=0.9\textwidth]{euler_different_timesteps.png}
     \caption{Relative energy error for the Euler integrator using different time-steps for a fixed total model simulation time.}
-    \label{fig:euler}
 \end{figure}
-As can be seen in figure~\ref{fig:euler} there is a spike around time 18 for time-step scale 1.
-This error has no apparent reason and in future time-steps it will decrease and stabilize around -1.
-Furthermore, it is interesting to see that with smaller time-step scales the total relative energy error increases faster.
-Meaning, that if we run the model for an infinite time the total error will explode and produce no meaning full results.
-This, while smaller time-step scales should increase the accuracy in terms of relative energy error.
-Therefore, we can conclude that the increase in accuracy does not outweighs the increase in iterations to complete the calculation.
 
@@
 \label{sec:res:taylor}
-Using the same energy metrics defined in section~\ref{XX} we can do the same simulation for the Taylor approximation.
-We expect that adding an extra term in the Taylor expansion would increase our accuracy.
-And in figure~\ref{fig:taylor} we can see that indeed slower increase in relative energy error than using bigger time-step scales.
-However, all the time-step scale converge to -1 after about 20\% of the total model simulation time.
-Also, we noticed that the 0.001 time-step scale converges faster to the -1 relative energy error than the 0.0001 time-step scale.
-Here, we can observe that decreasing the time-step by 10 does not increase the accuracy enough to compensate for the increase in number of iterations.
-The could be due to machine-precision, but requires further investigation to be sure.
-
-\begin{figure}
-    \includegraphics[width=0.9\textwidth]{jerk_different_timesteps_truncated.png}
-    \caption{Relative energy error for the Taylor integrator using different time-steps for a fixed total model simulation time.}
-    \label{fig:taylor}
-\end{figure}
 
@@
 \label{sec:res:leapfrog}
-Using the Leapfrog integrator, which is a slight modified Euler integrator, we should expect similar results.
-And as can be seen in figure~\ref{fig:leapfrog}.
-the relative energy error shows similar results as for the Euler approximation in figure~\ref{fig:euler}.
-However, the magnitude of the relative energy error is higher with the Leapfrog integrator at around time-step 18.
-This is probably due to the approximation of the in between time-steps which the Leapfrog integrator uses.
-Nevertheless, the results are relatively the same as the Euler approximation and does, at least in our simulation, lead to a decrease in relative energy error.
-
-\begin{figure}
-    \includegraphics[width=0.9\textwidth]{leapfrog_different_timesteps.png}
-    \caption{Relative energy error for the Leapfrog integrator using different time-steps for a fixed total model simulation time.}
-    \label{fig:leapfrog}
-\end{figure}
 
@@
 \label{sec:res:all}
-To compare all the models, i.e.
-Euler, Taylor and Leapfrog integrator, we setup up an experiment which uses the same time-step for all the models.
-The time-step is set at $0.01$ and figure~\ref{} shows the relative energy error for all models.
-Here we can clearly observe that the Taylor integrator gives the best results in terms of relative energy error.
-And as expected the Leapfrog and Euler integrator are approximately the same.
-However, the Taylor approximation comes with extra costs in computation time which will be discussed next.
-
-\begin{figure}
-    \includegraphics[width=0.9\textwidth]{constant_time.png}
-    \caption{Relative energy error for the Euler,Taylor and Leapfrog integrator using fixed time-steps for a fixed total model simulation time.}
-    \label{fig:leapfrog}
-\end{figure}
-
-\subsection{Complexity}
-\label{sec:res:complexity}
 

### 2014-10-25 article/article.tex

--- 2014-10-25 article/article.tex
@@
 
-\subsubsection{Sensitivity analysis}
-because we use approximations we have sensitive systems etc....
-perturbations in input variables and output variables. i.e. positions of other objects are incorrect.
-Delta step influences sensitivity?....
-
 \subsubsection{Model verification}

### 2014-10-25 article/article.tex
typos, close encounter sectie

@@
 To verify the accuracy of the simulation we can apply some fundamental conservation laws from physics to our simulation.
-We know based on empirical data that these laws seem to be true for the natural, so if our simulation is a faithful model of nature these laws should also be true for our simulation.
+We know based on empirical data that these laws seem to be true for the natural world, so if our simulation is a faithful model of nature these laws should also be true for our simulation.
 If they do not, the amount our simulation fails to conserve certain values that are supposed to be conserved can be used as a metric for the error in our simulation.\\\\
@@
 \end{equation}
-The total (mechanical) energy of system is the potential and kinetic energy of the system added together, $E = W+K$.
+The total (mechanical) energy of a system is the potential and kinetic energy of the system added together, $E = W+K$.
 This total energy should remain constant.
@@
 The important part in these simulations is the determination of the force each of the bodies experiences.
-In direct solvers, this is done the same for each body to body interactions and the total force on a particular body is a simple summation of the forces it experiences from every other body.
+In direct solvers, this is done the same for each body to body interaction and the total force on a particular body is a simple summation of the forces it experiences from every other body.
 \\
-To determine the force between two bodies due to gravity is a second order differential equation needs to be solved. Thus numerical integrators are the key component of direct solvers.
+To determine the force between two bodies due to gravity a second order differential equation needs to be solved. Thus numerical integrators are the key component of direct solvers.
 \subsubsection{Integrator Algorithms}
@@
 This is supposed to improve its energy conservation properties.
+\\\\
+\textbf{Handling Close Encounters (not of the Third Kind)}\\
+In a gravitational simulation with discrete timesteps strange things can happen if two bodies get really close to each other.
+If two bodies are really close to each other they may experience a very strong force interaction.
+In reality this would happen for a very short time before the bodies either collide or they move away from each other again and experience a decreased amount of force.\\
+The n-body simulation that has been studied uses a collisionless n-body model.
+This means that the first option for close encounters (collision) is disregarded.
+A simulation with discrete timesteps will assume that the force experienced at the start of a timestep by a body will be experienced by that body during the whole timestep.
+This is not necessarily true for two close bodies in collisionless system.
+The bodies would only experience the strong force for a much shorter time than the timestep actually last and thus the force interaction gets overestimated.
+\\
+One option to improve accuracy of close encounters is to have the simulation run with smaller timesteps.
+This would however increase computational load and be inefficient if most simulation steps don't actually have encounters close enough to need a shorter timestep.\\
+Instead a dynamic timestep system can be devised.
+By looking at each gravitational interaction in the system, the resulting speeds and accelerations between every pair of bodies due to force interaction can be derived.
+It can then be calculated how long it would take to cover the distance between the two bodies based on their force interaction.
+If the length of time is shorter than the default timestep length, the timestep for the simulation is lowered for the next simulation step.\\
+This lowering of the timestep is done globally, for the entire simulation.
+The simulation is slowed down based on the closest interaction between two bodies in the system.
+There are algorithms that lower the timestep per body, but those have not been looked at.
 \subsection{Implementation}

### 2014-10-25 article/article.tex
softening voor lengte tussen bodies, geen dynamic timestep voor leapfrog

@@
 The simulation is slowed down based on the closest interaction between two bodies in the system.
-There are algorithms that lower the timestep per body, but those have not been looked at.
+There are algorithms that lower the timestep per body, but those have not been looked at.\\
+The Euler and Jerk integrators can run in a optional dynamic timestep mode.
+The leapfrog because of its interleaving nature cannot legitimately be run in a dynamic timestep mode in the version implemented and has thus only been run in a fixed timestep mode.\\
+Additionally a softening factor has been added to the calculations of forces between two object.
+The calculation for acceleration (and jerk) divides by a power of the distance between two bodies, which means that if the distance gets close to zero the force gets multiplied by a very big number.
+To soften this and make sure that (as far as the force calculations go) can't get infinitely close a small $\epsilon$ is added to the distance between two bodies in these calculations.
 \subsection{Implementation}

### 2014-10-25 article/article.tex
implementation section

@@
 \subsection{Implementation}
-attachment
+The simulation was implemented in a C++11 application, with SDL2 and OpenGL for the visualisation.
+Calculations have been done with single precision floating point math, though the application is programmed in such a way that it is relatively easy to change this to double precision floating point math later.
+\\\\
+The implementation uses sequential (i.e. non-parallel) code to implement the studied n-body simulation algorithms. Data was written to textfiles and analyzed using R and Gnuplot.
 \section{Experimental Results}

### 2014-10-25 article/article.tex
proofread version

@@
 
-Although, it is not a new problem, because we have been trying to analyze how planets orbit each other in our own solar system for quite while, there is still room for improvement.
-The analytical solution for orbiting planets are only purposed for 2-body system, while in reality we have to cope with N-body systems.
+It is not a new problem, because we have been trying to analyze how planets orbit each other in our own solar system for quite while, there is still room for improvement.
+The analytical solution for orbiting planets  only exist for 2-body system, while in reality we have to cope with N-body systems.
 Approximation methods can be used only calculating is intractable by hand.
@@
 
-This research is aims to provide insight in the approximation methods used for N-body systems, i.e. how accurate is the simulation. 
+This research aims to provide insight with regards to the approximation methods used for N-body systems, i.e. how accurate is the simulation. 
 In section~\ref{sec:literature} the background of the analytical models used are explained and section~\ref{sec:methodology} introduces the algorithms as well a methods for benchmarking our approximation methods.
@@
 \label{sec:literature}
-In pursuance of calculating the trajectory of a body it is necessary to know which forces interact with our body.
+When calculating the trajectory of a body it is necessary to know which forces interact with our body.
 Newton's seconds law, i.e.
-the rate of change is equal to the net force on a object, is used to describe the force on our object.
+the acceleration is equal to the net force on a object, is used to describe the force on our object.
 And Newton's third law, i.e.
-action = - reaction, is used.
+every action has an equal and opposite reaction, is used.
 
@@
 Thus, if the object is accelerating there is a force on our object present.
-Knowing that the only possible object acting on our object are the other planets in the system we can modify the second law to derive the law of universal gravitation.
+Knowing that the only possible force acting on our object are originating from other bodies in the system we can modify the second law by equating the law of universal gravitation to it.
 Let the following equation be the force $F_p$ on object $p$ by object $c$ with mass $m_p,m_c$ respectively.
@@
     \label{eq:newtongravity}
-	a_p = G\frac{m_pm_c}{ m_pr^2}
+    \begin{split}
+	a_p = G\frac{m_pm_c}{ m_pr^2} \\
+	= G\frac{m_c}{r^2}
+    \end{split}
 \end{equation}.
@@
 \\
-\indent Assume we have a 2 body system called Pluto-Charon system. Figure~\ref{fig:plutocharon} illustrates this system at a fixed time point $t$
-\begin{figure*}
-	\label{fig:plutocharon}
-\end{figure*}
-with both $x,y$ coordinates in our 2D space.
-
-% Assume that we have an initial velocity vector $\overrightarrow{v_t}$ for the body t with speed in both x and y direction.
-% \begin{equation}
-% 	\overrightarrow{v_t} = \begin{bmatrix}
-% 								x_t \\
-% 								y_t
-% 							\end{bmatrix}
-% \end{equation}
-% For example we have a two body system with $p,c$ with an initial velocity vector, respectively, then
-% \begin{equation}
-% 	v_p=\begin{bmatrix} v_x^p \\ v_y^p \end{bmatrix}, v_c=\begin{bmatrix} v_x^c \\ v_y^c \end{bmatrix}
-% \end{equation}
+\indent Assume we have a 2 body system called Pluto-Charon system. 
 We assume that the whole system has a constant velocity, in our case we assume zero velocity, hence the next equation must be equal to zero
@@
 with $P$ as the total momentum and $M$ as the total mass of the system.
-We know that the total momentum is the sum of all forces acted up on and the mass is a constant.
+We know that the total momentum is the sum of all forces acted upon and the mass is a constant.
 Therefore, our zero velocity implies that the total momentum must be equal to zero.
@@
 \end{equation}
-As illustrated in figure~\ref{XXXX}(simple two body system with vectors drawn) we can see that each body has several forces acted upon.
-Our objective is the determine the path of body $p,c$ and as you can see we have a force vector $F_x$ pulling the body $X$ to the center and also our velocity vector $v_x$.
-The product of those vectors gives us the directional vector of our body $X$.
-Due to the fact that the two bodies are the only objects in the system acting on each other we can determine the radius and period of our movement around a (virtual) center of mass.
-We know for a fact that the two bodies must be the opposites of each other such that the total momentum is equal to zero.
-The center of mass can be calculated in reference of one of our bodies.
-(Bary center)
+Each body has several forces acted upon.
+Our objective is to determine the path of body $p,c$ and we have a force vector $F_x$ pulling a body $X$ to the other planet and also a velocity vector $v_x$ for each body.
+The product of those vectors gives us a directional vector of, for example, body $X$.
+The center of mass can be calculated relative to one of our bodies.
 \begin{equation}
@@
 \end{equation}
-With $r_m$ being the vector to the center of mass, the we can rewrite this to calculate $r_m$
+With $r_m$ being the vector to the center of mass, then we can rewrite this to calculate $r_m$
 \begin{equation}
@@
 \end{equation}
-
 If we want to know the force of body $p$ on body $c$ we can use Newton's law of gravitation, see equation~\ref{eq:newtongravity}.
@@
 The change of position i.e.
-distance of body $c$ at time $\delta t$ is given by
+distance of body $c$ at time $\Delta t$ is given by
 \begin{equation}
@@
 This can be obtained by taking the double integral of the acceleration.
-This can be solved analytically when only considering two bodies.
+This can be solved analytically when only considering two bodies \cite{frida2013}.
 If you consider $n \geq 3$ bodies it becomes unsolvable analytically.
@@
 We can derive the next position by adding the distance traveled in time $\Delta t$ to the old position at $x_t$, i.e.
-$x_{t+1} = x_t + v_t*\Delta t$ with $v_t$ as the speed at time $t$.
+$x_{t+1} = x_t + v_t\Delta t$ with $v_t$ as the speed at time $t$.
 Nonetheless, we do not know speed $v_t$ but this can be obtained by $v_t = v_t + a_t \Delta t$ and $a_t$ is known for our body.
@@
         v(t) =  \frac{v^{(0)}(t)}{0!}(x-t)^0 + \frac{v^{(1)}(t)}{1!}(x-t)^1 +\frac{v^{(2)}(t)}{2!}(x-t)^2\\
-        v(t) = v(0) + a(t) + j(t)
+        v(t) = v(t) + a(t)(x-t) + \frac{1}{2}j(t)(x-t)^2
     \end{split}
@@
         a(t) =  \frac{a^{(0)}(t)}{0!}(x-t)^0 + \frac{a^{(1)}(t)}{1!}(x-t)^1 \\
-        a(t) =  a(0) + j(t) 
+        a(t) =  a(t) + j(t)(x-t)
     \end{split}
@@
 we can calculate $j(t)$ by taking the derivative of equation~\ref{eq:newtongravity} with respect to $a$. Thus obtaining more accurate results for our $x(t), v(t)$ and $a(t)$.
-Also, error of the approximation can be calculated as follows:
 
@@
 \textbf{Conservation of Momentum}\\
-Momentum is defined as the mass of an object multiplied with its velocity vector. To get the momentum of a system of n bodies you can calculate the momentum of each body and sum them all together.
+Momentum is defined as the mass of an object multiplied with its velocity vector. To get the momentum of a system of N-bodies you can calculate the momentum of each body and sum them all together.
 \begin{equation}
@@
 On the one hand this means that a fundamental law is enforced with any algorithm that ends up with the same force with a body to body interaction for both bodies, with only the direction reversed for one of them.
-This will be true for the n-body algorithms covered in this paper.
-On the other this means that the conservation of momentum is not a useful error metric to check the accuracy of an algorithm.
-Hence it is not used in this paper to compare the accuracy of the studied algorithms.
-The measurements have been done, but show no real difference, even when it is clear that a simulation is doing things that are otherwise unrealistic and inaccurate.\\\\
+This will be true for the N-body algorithms covered in this paper.
+On the other hand this means that the conservation of momentum is not a useful error metric to check the accuracy of an algorithm.
+\\\\
 \textbf{Conservation of Energy}\\
@@
 In a system of moving point masses that only gravitationally interact (i.e.
-a n-body system) the energy in the system is the sum of the kinetic energies and potential energies of each of the masses in the system.
+an N-body system) the energy in the system is the sum of the kinetic energies and potential energies of each of the masses in the system.
 This sum should remain constant in a system that has no energy added or otherwise taken away from it.
@@
 \end{equation}
-The potential energy of a body is the amount of energy it experiences due to its interaction with a force field, which would be with the gravitational field in a n-body simulation.
+The potential energy of a body is the amount of energy it experiences due to its interaction with a force field, which would be with the gravitational field in a N-body simulation.
 The total potential energy of a system is the sum of the potential energy of all the bodies in the system.
@@
 In reality it would only experience this energy for a very short time.
-But in a simulation with discrete timesteps, the timestep may be longer than the time the body is supposed to experience the potential energy for.
-Thus it will create extra kinetic energy without that energy being based on potential energy that would be there in reality when it interact continuously with bodies rather than at discrete timesteps.
+But in a simulation with discrete time-steps, the time-step may be longer than the time the body is supposed to experience the potential energy for.
+Thus it will create extra kinetic energy without that energy being based on potential energy that would be there in reality when it interact continuously with bodies rather than at discrete time-steps.
 \\
@@
 \begin{equation}
+    \label{eq:energyerror}
     E_{error} = \frac{E_{current} - E_{initial}}{E_{current}}
@@
 \subsection{Algorithms}
-The n-body problem involves every body in a system interacting with every other body through gravity.
+The N-body problem involves every body in a system interacting with every other body through gravity.
 This means that there are $\mathcal{O}(n^2)$ interactions that need to be calculated.
-Very broadly n-body simulators fall into two categories.
+Very broadly N-body simulators fall into two categories.
 The first category contains direct solvers, which simulate the gravitational interactions between all bodies directly, e.g.
@@
 Alternatively, there are solvers that use a tree to represent the set of bodies in a hierarchical manner.
-These algorithms have each of the bodies interact with the tree instead of the full set of n-bodies.
+These algorithms have each of the bodies interact with the tree instead of the full set of N-bodies.
 This can lead to a computational complexity of $\mathcal{O}(n \log(n))$.
-The Barnes-Hut algorithm is a pioneering example of this class of n-body simulation algorithms \cite{barnes1986hierarchical}.\\
+The Barnes-Hut algorithm is a pioneering example of this class of N-body simulation algorithms \cite{barnes1986hierarchical}.\\
 We have opted to go for a direct simulator, because the implementation of such algorithms tends to be less complex.
@@
         x_{t+ 1} = x_t + v_t \Delta t + \frac{1}{2} a_t (\Delta t)^2 + \frac{1}{6} j_t (\Delta t)^3 \\
-        v_{t+ 1} = v_t a_t \Delta t + \frac{1}{2} j_t (\Delta t)^2 \\
+        v_{t+ 1} = v_t + a_t \Delta t + \frac{1}{2} j_t (\Delta t)^2 \\
         a_{t+ 1} = a_t + j_t \Delta t \\
@@
 \textbf{Leapfrog Integrator}\\
-The leapfrog integrator calculates the positions and velocities of the bodies in a simulated system at interleaved timesteps. 
+The leapfrog integrator calculates the positions and velocities of the bodies in a simulated system at interleaved time-steps. 
 \begin{equation}
@@
 \\\\
-\textbf{Handling Close Encounters (not of the Third Kind)}\\
-In a gravitational simulation with discrete timesteps strange things can happen if two bodies get really close to each other.
+\textbf{Handling Close Encounters }\\
+In a gravitational simulation with discrete time-steps strange things can happen if two bodies get really close to each other.
 If two bodies are really close to each other they may experience a very strong force interaction.
 In reality this would happen for a very short time before the bodies either collide or they move away from each other again and experience a decreased amount of force.\\
-The n-body simulation that has been studied uses a collisionless n-body model.
+The N-body simulation that has been studied uses a collisionless N-body model.
 This means that the first option for close encounters (collision) is disregarded.
-A simulation with discrete timesteps will assume that the force experienced at the start of a timestep by a body will be experienced by that body during the whole timestep.
+A simulation with discrete time-steps will assume that the force experienced at the start of a time-step by a body will be experienced by that body during the whole time-step.
 This is not necessarily true for two close bodies in collisionless system.
-The bodies would only experience the strong force for a much shorter time than the timestep actually last and thus the force interaction gets overestimated.
+The bodies would only experience the strong force for a much shorter time than the time-step actually last and thus the force interaction gets overestimated.
 \\
-One option to improve accuracy of close encounters is to have the simulation run with smaller timesteps.
-This would however increase computational load and be inefficient if most simulation steps don't actually have encounters close enough to need a shorter timestep.\\
-Instead a dynamic timestep system can be devised.
+One option to improve accuracy of close encounters is to have the simulation run with smaller time-steps.
+This would however increase computational load and be inefficient if most simulation steps don't actually have encounters close enough to need a shorter time-step.\\
+Instead a dynamic time-step system can be devised.
 By looking at each gravitational interaction in the system, the resulting speeds and accelerations between every pair of bodies due to force interaction can be derived.
 It can then be calculated how long it would take to cover the distance between the two bodies based on their force interaction.
-If the length of time is shorter than the default timestep length, the timestep for the simulation is lowered for the next simulation step.\\
-This lowering of the timestep is done globally, for the entire simulation.
+If the length of time is shorter than the default time-step length, the time-step for the simulation is lowered for the next simulation step.\\
+This lowering of the time-step is done globally, for the entire simulation.
 The simulation is slowed down based on the closest interaction between two bodies in the system.
-There are algorithms that lower the timestep per body, but those have not been looked at.\\
-The Euler and Jerk integrators can run in a optional dynamic timestep mode.
-The leapfrog because of its interleaving nature cannot legitimately be run in a dynamic timestep mode in the version implemented and has thus only been run in a fixed timestep mode.\\
+There are algorithms that lower the time-step per body, but those have not been looked at.\\
+The Euler and Jerk integrators can run in a optional dynamic time-step mode.
+The leapfrog because of its interleaving nature cannot legitimately be run in a dynamic time-step mode in the version implemented and has thus only been run in a fixed time-step mode.\\
 Additionally a softening factor has been added to the calculations of forces between two object.
@@
 \\\\
-The implementation uses sequential (i.e. non-parallel) code to implement the studied n-body simulation algorithms. Data was written to textfiles and analyzed using R and Gnuplot.
+The implementation uses sequential (i.e. non-parallel) code to implement the studied N-body simulation algorithms. Data was written to textfiles and analyzed using R and Gnuplot.
 \section{Experimental Results}
@@
 Using these methods allows us to compare the performance of each model.
-For all the energy comparison methods we used a fixed total \textit{model evaluation time}, i.e.
+For all the experiments with energy comparisons we used a fixed total \textit{model evaluation time}, i.e.
 we simulate 100 model minutes.
 Depending on the time-steps used in each method, this might take more or less iterations to calculate, meaning that the total simulation time increases.
-In
-section~\ref{sec:res:euler} compares the relative error of the Euler method using different time-steps.
-Also, section~\ref{sec:res:taylor} compares the improvement of adding an extra term in the Taylor expansion.
+Section~\ref{sec:res:euler} compares the relative error of the Euler method using different time-steps.
+Section~\ref{sec:res:taylor} compares the improvement of adding an extra term in the Taylor expansion.
 And, section~\ref{sec:res:leapfrog} shows the results using the Leapfrog integrator and all the models are compared in the last section~\ref{sec:res:all}.
@@
 To verify our model, we have defined the energy metrics.
-In this experiment the relative energy metric as defined in section~\ref{XXX} is used.
-In total we have compared four different time-steps to evaluate the performance effect of an increasing/decreasing time-step.
-The time-steps used are $0.001, 0.01, 0.1, 1$ and figure~\ref{fig:euler} shows a plot of the relative energy error for the \textit{model simulation time}.
+In this experiment the relative energy metric as defined in equation~\ref{eq:energyerror} is used.
+In total we have compared four different time-step scales to evaluate the performance effect of an increasing/decreasing time-step scale.
+The time-step scale used are $0.001, 0.01, 0.1, 1$ and figure~\ref{fig:euler} shows a plot of the relative energy error for the \textit{model simulation time}.
 \begin{figure}
     \includegraphics[width=0.9\textwidth]{euler_different_timesteps.png}
-    \caption{Relative energy error for the Euler integrator using different time-steps for a fixed total model simulation time.}
+    \caption{Relative energy error for the Euler integrator using different time-step scales for a fixed total model simulation time.}
     \label{fig:euler}
@@
 Meaning, that if we run the model for an infinite time the total error will explode and produce no meaning full results.
-This, while smaller time-step scales should increase the accuracy in terms of relative energy error.
-Therefore, we can conclude that the increase in accuracy does not outweighs the increase in iterations to complete the calculation.
+The expectation may have been that smaller time-step scales should increase the accuracy in terms of relative energy error.
+Therefore, we can conclude that the increase in accuracy per time-step does not outweighs the increase in iterations to complete the calculation.
 
-\subsection{Taylor}
+\subsection{Jerk Integrator}
 \label{sec:res:taylor}
-Using the same energy metrics defined in section~\ref{XX} we can do the same simulation for the Taylor approximation.
+Using the same energy metrics defined in equation~\ref{eq:energyerror} we can do the same simulation for the third order Taylor approximation.
 We expect that adding an extra term in the Taylor expansion would increase our accuracy.
@@
 Euler, Taylor and Leapfrog integrator, we setup up an experiment which uses the same time-step for all the models.
-The time-step is set at $0.01$ and figure~\ref{} shows the relative energy error for all models.
-Here we can clearly observe that the Taylor integrator gives the best results in terms of relative energy error.
+The time-step is set at $0.01$ and figure~\ref{fig:energy_comparison} shows the relative energy error for all models.
+Here we can clearly observe that the Taylor integrator with jerk gives the best results in terms of relative energy error.
 And as expected the Leapfrog and Euler integrator are approximately the same.
@@
     \caption{Relative energy error for the Euler,Taylor and Leapfrog integrator using fixed time-steps for a fixed total model simulation time.}
-    \label{fig:leapfrog}
+    \label{fig:energy_comparison}
 \end{figure}
@@
 \begin{figure}
-    \includegraphics[width=0.9\textwidth]{constant_time.png}
+    \includegraphics[width=0.9\textwidth]{time_bodies.png}
     \caption{Computation time to finish calculating the total model simulation time for a given number of bodies.}
@@
 \section{Future work}
-For future work we would like to recommend to try a time-step scale for each body, meaning that close encounters have smaller time-step scale than other bodies.
+For future work we would like to recommend to try a distinct time-step scale for each body, meaning that close encounters have smaller time-step scale than other bodies.
 This should lead to a smaller energy error while not slowing down the whole process.

### 2014-10-25 code/cppsimulator/integrator.cpp
expanded readme, removed comments

@@ -18,10 +18,7 @@ void Integrator::jerkIntegrator(std::vector<Particle>& particles, bool dynamic_t
     //run simulation
     for(auto& particle: particles)
     {
-        //particle.calculateAcceleration(particles, timestep);
         particle._jerk = Vec3D();
-        //_acceleration = Vec3D();
-
         particle._potentialEnergy = 0;
         const float G = 0.01;
 
@@ -63,7 +60,7 @@ void Integrator::jerkIntegrator(std::vector<Particle>& particles, bool dynamic_t
         particle._position = particle._position +  particle._velocity * timestep + particle._acceleration*timestep*timestep/2 + (particle._jerk*timestep*timestep*timestep)/6;
         particle._velocity = particle._velocity + particle._acceleration * timestep + (particle._jerk*timestep*timestep)/2;
 
-        const int historySize = 20;
+        const int historySize = 40;
         particle._positionHistory.push_back(particle._position);
         if (particle._positionHistory.size() > historySize)
         {
@@ -114,7 +111,7 @@ void Integrator::naiveIntegrator(std::vector<Particle>& particles, bool dynamic_
         particle._position = particle._position +  particle._velocity * timestep;
         particle._velocity = particle._velocity + particle._acceleration * timestep;
 
-        const int historySize = 20;
+        const int historySize = 40;
         particle._positionHistory.push_back(particle._position);
         if (particle._positionHistory.size() > historySize)
         {
@@ -166,7 +163,7 @@ void Integrator::leapfrogIntegrator(std::vector<Particle>& particles, bool dynam
         particle._position = particle._position +  particle._velocity * timestep + particle._prevAcceleration * 0.5 * timestep * timestep;
         particle._velocity = particle._velocity + (particle._acceleration + particle._prevAcceleration) * 0.5 * timestep;
 
-        const int historySize = 20;
+        const int historySize = 40;
         particle._positionHistory.push_back(particle._position);
         if (particle._positionHistory.size() > historySize)
         {

### 2014-10-25 code/cppsimulator/main.cpp

@@ -204,13 +204,6 @@ int main(int argc, char* argv[])
             float relativeEnergyError = (initialTotalEnergy - totalEnergy)/
                 totalEnergy;
 
-            // std::cout << "total momentum: (" << totalMomentum.length() << ") " ;
-            // totalMomentum.print();
-            // std::cout << "total kinetic energy: " << totalKineticEnergy << std::endl;
-            // std::cout << "total potential energy: " << totalPotentialEnergy << std::endl;
-            // std::cout << "total energy: " << totalEnergy << std::endl;
-            // std::cout << "relative energy error: " << relativeEnergyError << std::endl;
-
             if(!outfile.is_open()) {
                 std::cerr << "Couldn't open 'log_file.txt'" << std::endl;
                 return -1;

### 2014-10-25 deliverable/code/integrator.cpp
Adds deliverable

@@ -1,0 +1,173 @@
+#include "integrator.h"
+#include "particle.h"
+#include <cmath>
+#include <iostream>
+#include <algorithm>
+
+#define GLOBAL_MINIMUM_TIMESTEP 0.1
+#define GLOBAL_MAXIMUM_TIMESTEP 0.001
+
+Integrator::Integrator() :
+    timestep(0.1)
+{
+}
+
+void Integrator::jerkIntegrator(std::vector<Particle>& particles, bool dynamic_time, float initial_timestep)
+{
+    float globalMin;
+    //run simulation
+    for(auto& particle: particles)
+    {
+        particle._jerk = Vec3D();
+        particle._potentialEnergy = 0;
+        const float G = 0.01;
+
+        for (const auto& p: particles)
+        {
+            if (&p == &particle)
+                continue;
+
+            Vec3D positionDelta = (p._position - particle._position);
+            Vec3D velocityDelta = p._velocity - particle._velocity;
+            float l = positionDelta.length() + 0.1;
+            particle._potentialEnergy += (particle._mass * p.mass() * G) / (l*l);
+            Vec3D jerkDelta = (((velocityDelta)/(l*l*l)) - 
+                    ((positionDelta*(positionDelta*velocityDelta))/(l*l*l*l*l))*3)*p.mass();
+            particle._jerk += jerkDelta;
+
+            Vec3D accelerationDelta = jerkDelta * timestep;
+
+            float minV = positionDelta.length() / velocityDelta.length();
+            float minA = sqrt(positionDelta.length() / accelerationDelta.length());
+            float minJ = powf(positionDelta.length() / jerkDelta.length(), 1.0/3);
+
+            globalMin = minV < globalMin ? minV : globalMin;
+            globalMin = minA < globalMin ? minA : globalMin;
+            globalMin = minJ < globalMin ? minJ : globalMin;
+    }
+
+    //Taylor expansion around the current point
+    particle._acceleration = particle._acceleration + particle._jerk*timestep;
+    
+    }
+
+    timestep = (dynamic_time)?fmin(globalMin, GLOBAL_MINIMUM_TIMESTEP):initial_timestep;
+    timestep = (dynamic_time)?fmin(timestep, GLOBAL_MAXIMUM_TIMESTEP):initial_timestep;
+
+    for(auto& particle: particles)
+    {
+        //Taylor expansion around the current point
[…]

### 2014-10-25 deliverable/code/integrator.h

@@ -1,0 +1,20 @@
+#ifndef INTEGRATOR_H_
+#define INTEGRATOR_H_
+
+#include <vector>
+
+class Particle;
+
+class Integrator
+{
+    public:
+        Integrator();
+        void naiveIntegrator(std::vector<Particle>& particles, bool dynamic_timestep, float initial_timestep);
+        void leapfrogIntegrator(std::vector<Particle>& particles, bool dynamic_timestep, float initial_timestep);
+        void jerkIntegrator(std::vector<Particle>& particles, bool dynamic_timestep, float initial_timestep);
+        float getTimestep() {return timestep;}
+    private:
+        float timestep;
+};
+
+#endif // INTEGRATOR_H_

### 2014-10-25 deliverable/code/main.cpp

@@ -1,0 +1,283 @@
+#include <iostream>
+#include <vector>
+#include <OpenGL/gl.h>
+#include <OpenGL/glu.h>
+#include "SDL.h"
+#include "particle.h"
+#include "integrator.h"
+#include "timer.h"
+#include <fstream>
+#include <cstdlib>
+
+
+SDL_Window* initSDLAndOpenGL(int width, int height, bool vsync, bool fullscreen)
+{
+    SDL_Init(SDL_INIT_VIDEO);
+    SDL_GL_SetAttribute( SDL_GL_RED_SIZE, 8 );
+    SDL_GL_SetAttribute( SDL_GL_GREEN_SIZE, 8 );
+    SDL_GL_SetAttribute( SDL_GL_BLUE_SIZE, 8 );
+    SDL_GL_SetAttribute( SDL_GL_DEPTH_SIZE, 32 );
+    SDL_GL_SetAttribute( SDL_GL_DOUBLEBUFFER, 1 );
+    SDL_GL_SetAttribute( SDL_GL_MULTISAMPLEBUFFERS , 1 );
+    SDL_GL_SetAttribute( SDL_GL_MULTISAMPLESAMPLES , 4 );
+    SDL_Window* screen;
+    if (fullscreen)
+    {
+        screen = SDL_CreateWindow("N-Body simulator",
+                SDL_WINDOWPOS_UNDEFINED,
+                SDL_WINDOWPOS_UNDEFINED,
+                width, height,
+                SDL_WINDOW_OPENGL | SDL_WINDOW_FULLSCREEN_DESKTOP);
+    }
+    else
+    {
+        screen = SDL_CreateWindow("N-Body simulator",
+                SDL_WINDOWPOS_UNDEFINED,
+                SDL_WINDOWPOS_UNDEFINED,
+                width, height,
+                SDL_WINDOW_OPENGL);
+    }
+
+    SDL_GLContext glcontext = SDL_GL_CreateContext(screen);
+
+    if (SDL_GL_SetSwapInterval(vsync? 1 : 0) == -1) {
+        std::cout << "could not set desired vsync mode: " << vsync << 
+            std::endl;
+    }
+    glClearColor(0.1, 0.1, 0.15, 0);
+
+    //glClearDepth(1);
+    //glDepthFunc(GL_LESS);
+
+    glViewport(0, 0, width, height);
+
+    glMatrixMode(GL_PROJECTION);
+    glLoadIdentity();
+    gluPerspective(90, float(width)/height, 0.1, 100.0);
[…]

### 2014-10-25 deliverable/code/particle.cpp

@@ -1,0 +1,51 @@
+#include "particle.h"
+
+#include <iostream>
+#include <OpenGL/gl.h>
+
+Particle::Particle() : Particle(1, 0.01)
+{
+}
+
+Particle::Particle(float positionSize, float velocitySize)
+{
+    _position = Vec3D::createRandom(-positionSize, positionSize);
+    _velocity = Vec3D::createRandom(-velocitySize, velocitySize);
+    _acceleration = Vec3D();
+    _mass = 1;
+}
+
+
+
+void Particle::render()
+{
+    float redness = _potentialEnergy * 20;
+    if (redness > 1)
+        redness = 1;
+    glColor3f(1,1-redness,1-redness);
+    glBegin(GL_POINTS);
+    glVertex3f(_position.x(), _position.y(), _position.z());
+    glEnd();
+
+    glBegin(GL_LINE_STRIP);
+    int i = 0;
+    for (auto position: _positionHistory)
+    {
+        float c = float(i) / _positionHistory.size();
+        glColor3f(c,c,c);
+        i++;
+        glVertex3f(position.x(), position.y(), position.z());
+    }
+    glEnd();
+}
+
+Vec3D Particle::momentum()
+{
+    return _velocity * _mass;
+}
+
+float Particle::kineticEnergy()
+{
+    float v = _velocity.length();
+    return 0.5 * _mass * v * v;
+}

### 2014-10-25 deliverable/code/particle.h

@@ -1,0 +1,30 @@
+#ifndef PARTICLE_H_
+#define PARTICLE_H_
+
+#include <vector>
+#include <deque>
+#include "vec.h"
+
+class Particle
+{
+    public:
+        Particle();
+        Particle(float positionSize, float velocitySize);
+        void render();
+        Vec3D _position;
+        Vec3D momentum();
+        float kineticEnergy();
+        float mass() const {return _mass;}
+        float potentialEnergy() {return _potentialEnergy;}
+        friend class Integrator;
+    private:
+        std::deque<Vec3D> _positionHistory;
+        float _potentialEnergy;
+        Vec3D _velocity;
+        Vec3D _acceleration;
+        Vec3D _prevAcceleration;
+        float _mass;
+        Vec3D _jerk;
+};
+
+#endif // PARTICLE_H_

### 2014-10-25 deliverable/code/runsim.sh

@@ -1,0 +1,46 @@
+#!/bin/bash          
+rm log_file.txt
+#./sim leapfrog norender
+#mv log_file.txt leapfrog.txt
+#./sim jerk norender
+#mv log_file.txt jerk.txt
+#./sim jerk dynamic norender
+#mv log_file.txt jerk_dynamic.txt
+#./sim naive norender
+#mv log_file.txt naive.txt
+#./sim naive dynamic norender
+#mv log_file.txt naive_dynamic.txt
+#
+#
+#./sim naive norender timestep=0.001
+#mv log_file.txt naive.0_001.txt
+#./sim naive norender timestep=0.01
+#mv log_file.txt naive.0_01.txt
+#./sim naive norender timestep=0.1
+#mv log_file.txt naive.0_1.txt
+#./sim naive norender timestep=1
+#mv log_file.txt naive.1.txt
+#
+#./sim leapfrog norender timestep=0.001
+#mv log_file.txt leapfrog.0_001.txt
+#./sim leapfrog norender timestep=0.01
+#mv log_file.txt leapfrog.0_01.txt
+#./sim leapfrog norender timestep=0.1
+#mv log_file.txt leapfrog.0_1.txt
+#./sim leapfrog norender timestep=1
+#mv log_file.txt leapfrog.1.txt
+#
+#./sim jerk norender timestep=0.001
+#mv log_file.txt jerk.0_001.txt
+#./sim jerk norender timestep=0.01
+#mv log_file.txt jerk.0_01.txt
+#./sim jerk norender timestep=0.1
+#mv log_file.txt jerk.0_1.txt
+#./sim jerk norender timestep=1
+#mv log_file.txt jerk.1.txt
+
+
+#./sim naive norender timetrial=11 timestep=0.1 bodies=2
+#mv log_file.txt naive_time_bodies.txt
+./sim jerk norender timetrial=11 timestep=0.1 bodies=2
+mv log_file.txt jerk_time_bodies.txt

### 2014-10-25 deliverable/code/timer.cpp

@@ -1,0 +1,58 @@
+#include "SDL.h"
+#include "timer.h"
+
+Timer::Timer():
+    _totalTicks(0), _currentCache(0)
+{
+    for (int i = 0; i < CACHE_SIZE; i++)
+        _tickCache[i] = 1000/60;
+}
+
+void Timer::start()
+{
+  _ticks = SDL_GetTicks();
+}
+
+double Timer::reset()
+{
+  int previousTicks = _ticks;
+  start();
+  int difference = _ticks - previousTicks;
+
+
+  _totalTicks += difference;
+  _tickCache[_currentCache] = difference;
+  _currentCache = (_currentCache+1)%CACHE_SIZE;
+  int totalDifference = 0;
+  for (int i = 0; i < CACHE_SIZE; i++)
+      totalDifference += _tickCache[i];
+  return difference / (1000.0);
+}
+
+double Timer::total()
+{
+  return _totalTicks / 1000.0;
+}
+
+double Timer::stop()
+{
+  int totalTicks = _totalTicks;
+  _totalTicks = 0;
+  return totalTicks / 1000.0;
+}
+
+bool Timer::wait(int seconds)
+{
+    int current = SDL_GetTicks();
+    return (current - _ticks) < (seconds * 1000);
+}
+
+std::vector<int> Timer::getTickHistory()
+{
+    std::vector<int> history;
+    for (int i = 0; i < CACHE_SIZE; i++)
+    {
+        history.push_back(_tickCache[(i + _currentCache)%CACHE_SIZE]);
+    }
+    return history;
+}

### 2014-10-25 deliverable/code/vec.cpp

@@ -1,0 +1,78 @@
+#include "vec.h"
+
+#include <cstdlib>
+#include <cmath>
+#include <iostream>
+
+Vec3D::Vec3D():
+   _x(0), _y(0), _z(0) 
+{
+}
+
+Vec3D::Vec3D(float x, float y, float z):
+   _x(x), _y(y), _z(z) 
+{
+}
+
+Vec3D Vec3D::createRandom(Vec3D min, Vec3D max)
+{
+    float x = ((max._x-min._x)*((float)rand()/RAND_MAX))+min._x;
+    float y = ((max._y-min._y)*((float)rand()/RAND_MAX))+min._y;
+    float z = ((max._z-min._z)*((float)rand()/RAND_MAX))+min._z;
+    return Vec3D(x, y, z);
+}
+
+Vec3D Vec3D::operator+(const Vec3D& vec) const
+{
+    float x = _x + vec._x;
+    float y = _y + vec._y;
+    float z = _z + vec._z;
+    return Vec3D(x, y, z);
+}
+
+Vec3D Vec3D::operator-(const Vec3D& vec) const
+{
+    float x = _x - vec._x;
+    float y = _y - vec._y;
+    float z = _z - vec._z;
+    return Vec3D(x, y, z);
+}
+
+float Vec3D::length()
+{
+    return sqrt(_x*_x + _y*_y + _z*_z);
+}
+
+Vec3D Vec3D::createRandom(float min, float max)
+{
+    return createRandom(Vec3D(min, min, min), Vec3D(max, max, max));
+}
+
+
+Vec3D Vec3D::operator*(float scalar)
+{
+    return Vec3D(_x*scalar, _y*scalar, _z*scalar);
+}
+
+float Vec3D::operator*(const Vec3D& vec) const
+{
+    return (_x*vec.x() + _y*vec.y() + _z*vec.z());
+}
+
+Vec3D Vec3D::operator/(float scalar)
+{
+    return Vec3D(_x/scalar, _y/scalar, _z/scalar);
+}
+
+Vec3D& Vec3D::operator+=(const Vec3D& vec)
+{
+    _x += vec._x;
+    _y += vec._y;
+    _z += vec._z;
+    return (*this);
+}
+
+void Vec3D::print()
+{
+    std::cout << _x << " " << _y << " " << _z << std::endl;
+}

### 2014-10-25 deliverable/code/vec.h

@@ -1,0 +1,27 @@
+#ifndef VEC_H_
+#define VEC_H_
+
+class Vec3D
+{
+    public:
+        Vec3D(float x, float y, float z);
+        Vec3D();
+        static Vec3D createRandom(Vec3D min, Vec3D max);
+        static Vec3D createRandom(float min, float max);
+        Vec3D operator*(float scalar);
+        Vec3D operator/(float scalar);
+        float operator*(const Vec3D& vec) const;
+        Vec3D operator+(const Vec3D& vec) const;
+        Vec3D operator-(const Vec3D& vec) const;
+        float length();
+        void print();
+
+        Vec3D& operator+=(const Vec3D& vec);
+        float x() const {return _x;}
+        float y() const {return _y;}
+        float z() const {return _z;}
+    private:
+        float _x, _y, _z;
+};
+
+#endif // VEC_H_
